---
title: "Patient data analysis"
output: pdf_document
bibliography: LeafRank.bib
---

Script description:

This script aims to streamline HGSOC patient data analysis for LeafRank.
The backbone for this script is plotting the main figures in manuscript, i.e.
Figure 5(Patient base), Figure 6(Patient aneuploidy pattern), and Figure 7(In silico WGD robustness).

## Initialization

```{r}
library('rhdf5')
library('data.table')
library('reticulate')
library("ape")
library("tseries")
library("tibble")
library("ggplot2")
library("stringr")
library(circlize)
library(Cairo)
library("ggtreeExtra")
library("ggpubr")
library("MASS")
library("cluster")
library(uwot)
library(RColorBrewer)
library(doParallel)
library(foreach)
library(here)
library(ggtree)
library(mclust)
library(dbscan)
library(ComplexHeatmap)
library("R.matlab")
library("tidyr")
library(dplyr)
library(castor)
library(ggnewscale)


source('../R/tree_utils.R')
source('../R/fitness_inference.R')
source('../R/solve_density.R')
source('../R/message_passing.R')

# hg38 <- read.table("ith.Fitness/R/cytoband_hg38.txt", header = FALSE, sep = "\t")
# 
# colnames(hg38) <- c("Chr", "Start bin", "End bin", "Position")


hg19 <- read.table("cytoband_hg19.txt", header = FALSE, sep = "\t")  ## Read the cytoband here.

colnames(hg19) <- c("Chr", "Start bin", "End bin", "Position")


OV_gene <- data.frame(
  name = c("CCNE1", "MECOM", "MYC","KRAS", "RB1", "NF1", "PTEN"),
  type = c("OG","OG","OG","OG", "TSG","TSG","TSG"),
  chr  = c("19","3","8","12", "13", "17", "10"),
  location = c("q12", "q26.2", "q24.21","p12.1", "q14.2", "q11.2", "q23.31")
)

start <- c()
end   <- c()
for (i in 1:dim(OV_gene)[1]){
  chr <- paste0("chr",as.character(OV_gene$chr[i]))
  loc <- OV_gene$location[i]
  start_i <- hg19$`Start bin`[hg19$Chr == chr & hg19$Position == loc]
  end_i   <- hg19$`End bin`[hg19$Chr == chr & hg19$Position == loc]
  start <- c(start, start_i)
  end   <- c(end, end_i)
}

OV_gene$start <- start
OV_gene$end   <- end
OV_gene$start_idx <- rep(0,length(start))
OV_gene$end_idx <- rep(0,length(end))

```

## Figure 5(Patient base)

We first examine inference results from OV truncal WGD patient data (OV-003,OV-008,OV-014,OV-049,OV-068,OV-082,OV-087,OV-105,OV-118,OV-129).
This Figure includes: Panel A: Fitness plot for OV-105 Panel B: leave-m-out validation results for all 10 patient.

### Figure 5 panel A

```{r}
OV_num <- "105"
dir_name <- paste0("Results/patient results/Truncal/OV-",OV_num,"/") ##DATA
results_file <- paste0(dir_name,"pFitness_ultra.rds")           ##DATA
Results <- readRDS(results_file)
tree <- Results$phylo
meanFitness <- Results$meanFitness[1:length(tree$tip.label)]
mp <- Results$marginalProb[1:length(tree$tip.label),]
u_tree  <- readRDS(paste0(dir_name,"ultra_tree.rds"))
hc_tree <- as.hclust(u_tree)
b_rates <- 10*(1.1*1.1^(0:15))

rownames(mp) <- tree$tip.label
colnames(mp) <- c("Type0", "Type1","Type2","Type3","Type4","Type5","Type6","Type7","Type8","Type9","Type10","Type11","Type12","Type13","Type14","Type15")
df <- as.data.frame(mp)
df$label <- rownames(mp)


prob_long <- df %>%
  pivot_longer(cols = -label, names_to = "Type", values_to = "Probability") 

prob_long$xpos <- rep(b_rates, times = length(tree$tip.label))

mf_df <- data.frame(
  label = tree$tip.label,
  fitness = meanFitness
)

cluster_h <- tree$tip.label[meanFitness > 2.8]
cluster_m <- tree$tip.label[meanFitness > 2 ]
cluster_l <- tree$tip.label[meanFitness >1.5]

MRCA_h    <- getMRCA(tree,cluster_h)
MRCA_m    <- getMRCA(tree,cluster_m)
MRCA_l    <- getMRCA(tree,cluster_l)

Branch_h <- which(tree$edge[,2] == MRCA_h)
Branch_m <- which(tree$edge[,2] == MRCA_m)
Branch_l <- which(tree$edge[,2] == MRCA_l)

parent_h <- tree$edge[Branch_h,1]
parent_m <- tree$edge[Branch_m,1]
parent_l <- tree$edge[Branch_l,1]





p <- ggtree(tree, size = 0.3)+ geom_point2(aes(subset = node == parent_h),size = 4, shape = 24, fill = "#A1798F") +
  geom_point2(aes(subset = node == parent_m),size = 4, shape = 24, fill = "#A1798F") + 
  geom_point2(aes(subset = node == parent_l),size = 4, shape = 24, fill = "#A1798F")

p <- p  + geom_facet(
        panel = "Fitness",
        data = mf_df,
        geom = geom_col,
        mapping = aes(x = fitness, fill = fitness),
        orientation = "y",
        stat = "identity",
        width = 0.5
        
      ) +
    scale_fill_gradient(low = "#93C477", high = "#1F78B4") + guides(fill = guide_legend(order = 1), color = guide_legend(order = 2)) 




p <- p + new_scale_fill()





```

### Figure 5 panel B

```{r}
Valid_case <- c("003","008", "014", "049", "068", "082","087", "105", "118", "129")
#Valid_case <- c("008")
corr_df <- data.frame(
  idx_num = 1:50
)
ori_fit_stat <- data.frame(
  stat = c( "Mean", "SD","SD_u","Cell number","corr_ultra_poi","corr_ultra_nb","corr_poi_nb","Colless","Colless_normalized","Sackin","Blum","max_fit","max_fit_u")
)

for (case in Valid_case){
  dir_name <- paste0("Results/patient results/Truncal/OV-",case,"/")         ## DATA
  Results_ori <- readRDS(paste0(dir_name,"pFitness_ori.rds"))           ## DATA
  tree_ori <- Results_ori$phylo
  mf_ori <- Results_ori$meanFitness[1:length(tree_ori$tip.label)]
  Results_ultra <- readRDS(paste0(dir_name,"pFitness_ultra.rds"))       ## DATA
  tree_ultra <- Results_ultra$phylo
  mf_ultra <- Results_ultra$meanFitness[1:length(tree_ultra$tip.label)]
  Results_ultra_nb <- readRDS(paste0(dir_name,"pFitness_ultra_nb.rds")) ## DATA
  tree_ultra_nb <- Results_ultra_nb$phylo
  mf_ultra_nb <- Results_ultra_nb$meanFitness[1:length(tree_ultra$tip.label)]
  corr_hist <- c()
  for (i in 1:50){
    Results_sub <- readRDS(paste0(dir_name,"Subsampling/pFitness_subtree_",as.character(i),".rds"))                                                                  ## DATA
    tree_sub <- Results_sub$phylo
    mf_sub   <- Results_sub$meanFitness[1:length(tree_sub$tip.label)]
    idx_i <- match(tree_sub$tip.label, tree_ori$tip.label)
    mf_ori_i <- mf_ori[idx_i]
    corr_sub <- cor(mf_ori_i,mf_sub,method = "spearman")
    corr_hist <- c(corr_hist,corr_sub)
  }
  colname <- paste0("OV-",case)
  
    ori_fit_stat[[colname]] <- c(mean(mf_ori), sd(mf_ori),sd(mf_ultra), length(tree_ori$tip.label), cor(mf_ori, mf_ultra, method = "spearman"), cor(mf_ori,mf_ultra_nb, method = "spearman"), cor(mf_ultra_nb,mf_ultra, method = "spearman"),tree_imbalance(tree_ori,'Colless'),tree_imbalance(tree_ori,'Colless_normalized'),tree_imbalance(tree_ori,'Sackin'),tree_imbalance(tree_ori,'Blum'),max(mf_ori),max(mf_ultra))
  corr_df[[colname]] <- corr_hist
}


corr_df_long <- corr_df|>
  pivot_longer(
    cols = -idx_num,
    names_to = "Case",
    values_to = "Accuracy"
  )

p <- ggplot(corr_df_long, aes(x = Case, y = Accuracy, fill = Case)) +
  geom_violin(trim = FALSE, alpha = 0.5) +
  geom_boxplot(width = 0.1, outlier.shape = NA) +
  theme_bw()


#ggsave("Andrew2025/Results/Figures/Cross-validation.svg", p, width = 16, height = 8)

```

### Suppplementary plots

The following scripts are used to generate the fitness inference from all ten patients.
As an illustration, we include the examination of re-scaled ultrametric tree.

We first examine through rootogram for the ultrametric tree reconstructed based on Poisson distribution and Negative binomial distribution.
Neither Poisson nor Negative binomial has dominant performance here.

#### (Figure s3 and s4) Plot Truncal rootogram (Poisson)

```{r}
OV_num <- "118"
OV_class <- "Truncal"
dir_name <- paste0("Results/patient results/", OV_class,"/OV-",OV_num,"/")

tree_ori <- readRDS(paste0(dir_name,"tree_ori.rds"))
tree_ultra <- readRDS(paste0(dir_name,"ultra_tree.rds"))

y <- tree_ori$edge.length
t <- tree_ultra$edge.length

WGD_file <- paste0(dir_name,"ultra-tree.csv")
WGD_Results <- read.csv(WGD_file, header = FALSE)
WGD_rates <- WGD_Results[2,]

tau <- WGD_rates*t
n   <- length(y)

k_max <- max(y)
k <- 0:k_max
O <- tabulate(y+1, nbins = k_max+1)
E <- sapply(k, function(kk){
  sum(dpois(kk,lambda = tau))
})


chisq_test <- chisq.test(x = O, p = E/sum(E))

# barplot(
#   sqrt(O) - sqrt(E),
#   names.arg = k,
#   ylab = expression(sqrt(O[k]) - sqrt(E[k])),
#   xlab = "Count k",
#   main = "Poisson Mixture Rootogram",
#   col = "steelblue"
# )
# abline(h = 0, lwd = 2)


sqrt_exp <- sqrt(E)
sqrt_obs <- sqrt(O)

obs_stat <- 2 * sum(
  ifelse(y == 0,
         0,
         y * log(y / tau)) -
    (y - tau)
)

B <- 1000
sim_stat <- numeric(B)
for (b in 1:B) {
  sim_branch <- rpois(length(tau), tau)
  
  sim_stat[b] <- 2 * sum(
    ifelse(sim_branch == 0,
           0,
           sim_branch * log(sim_branch / tau)) -
      (sim_branch -  tau)
  )
}

p_sim <- mean(sim_stat >= obs_stat)
p_sim


#svg(paste0("Andrew2025/Results/Figures/Truncal_Fitness/OV-",OV_num,"-rooto.svg"), width = 6, height = 4)
plot(k, sqrt_exp,
     type = "n",
     ylim = range(c(sqrt_exp, sqrt_exp - sqrt_obs)),
     xlab = "Count k",
     ylab = expression(sqrt(E[k]) - sqrt(O[k])),
     main = "Hanging rootogram")

# expected curve
lines(k, sqrt_exp, lwd = 2)

# hanging bars
segments(x0 = k,
         y0 = sqrt_exp,
         x1 = k,
         y1 = sqrt_exp - sqrt_obs,
         lwd = 6)

abline(h = 0, lty = 2)
#dev.off()

p <- recordPlot()




signs <- sign(sqrt_exp - sqrt_obs)
sign_test <- runs.test(as.factor(signs))


```

#### (Figure s3 and s4) Plot Truncal rootogram (Negative binomial)

```{r}
OV_num <- "008"
OV_class <- "Truncal"
dir_name <- paste0("Results/patient results/",OV_class,"/OV-",OV_num,"/")

tree_ori <- readRDS(paste0(dir_name,"tree_ori.rds"))
tree_ultra <- readRDS(paste0(dir_name,"ultra_tree_nb.rds"))

y <- tree_ori$edge.length
t <- tree_ultra$edge.length

WGD_file <- paste0("Results/patient results/",OV_class,"/OV-",OV_num,"/ultra-tree_nb.csv")
WGD_Results <- read.csv(WGD_file, header = FALSE)
WGD_rates <- WGD_Results[2,]
r         <- WGD_Results[dim(WGD_Results)[1],]

p <- r/(r + WGD_rates*t)

k_max <- max(y)
k <- 0:k_max
O <- tabulate(y+1, nbins = k_max+1)
E <- sapply(k, function(kk){
  sum(dnbinom(kk,size = r, prob = p))
})

chisq_test <- chisq.test(x = O, p = E/sum(E))

# barplot(
#   sqrt(O) - sqrt(E),
#   names.arg = k,
#   ylab = expression(sqrt(O[k]) - sqrt(E[k])),
#   xlab = "Count k",
#   main = "Poisson Mixture Rootogram",
#   col = "steelblue"
# )
# abline(h = 0, lwd = 2)


sqrt_exp <- sqrt(E)
sqrt_obs <- sqrt(O)
#svg(paste0("Andrew2025/Results/Figures/Truncal_Fitness/OV-",OV_num,"-rooto-nb.svg"), width = 6, height = 4)
plot(k, sqrt_exp,
     type = "n",
     ylim = range(c(sqrt_exp, sqrt_exp - sqrt_obs)),
     xlab = "Count k",
     ylab = expression(sqrt(E[k]) - sqrt(O[k])),
     main = "Hanging rootogram")

# expected curve
lines(k, sqrt_exp, lwd = 2)

# hanging bars
segments(x0 = k,
         y0 = sqrt_exp,
         x1 = k,
         y1 = sqrt_exp - sqrt_obs,
         lwd = 6)

abline(h = 0, lty = 2)
#dev.off()

p <- recordPlot()



signs <- sign(sqrt_exp - sqrt_obs)
sign_test <- runs.test(as.factor(signs))
```

We then plot all fitness ranking inference from reconstructed ultrametric tree ("p_u").
The spearman correlation between these two type of inference are plotted in title of ultrametric fitness plots.
We also examine the marginal probability density plot as "mp_ht".
To reproduce all patients' fitness inference, one can ajust index `i` correspondingly.



#### (Figure s3 and s4) Plot the other Truncal fitness prediction and clustering them by fitness 

```{r}

Num_cluster <- c(3,2,3,3,2,3,2,2,2,4)#c(2,2,2,2,2,2,2,2,2,2)#HL: c(3,2,3,3,2,3,2,3,2,4), LH: c(4,2,5,3,3,4,3,5,4,4)
OV_case <- c("003","008","014", "049","068","082","087","105","118","129")
i <- 6

OV_num <- OV_case[i]
dir_name <- paste0("Results/patient results/Truncal/OV-",OV_num,"/")
results_file <- paste0(dir_name,"pFitness_ori.rds")
Results <- readRDS(results_file)
tree <- Results$phylo
meanFitness <- Results$meanFitness[1:length(tree$tip.label)]
mp <- Results$marginalProb[1:length(tree$tip.label),]
b_rates <- 1.1*1.1^(0:15)
colnames(mp) <- c(as.character(round(b_rates - 1,2)))
u_tree <- readRDS(paste0(dir_name,"ultra_tree.rds"))
hc_tree <- as.hclust(u_tree)




  
ring <- tibble(
  label = tree$tip.label,
  fitness = meanFitness
)
  


  
  p <- ggtree(tree, size = 0.3)+ ggtitle(paste0("OV-",OV_num," Fitness")) +
    labs(fill = "Fitness") +
    theme(legend.position = "bottom", 
          plot.title = element_text(size = 12, hjust = 0.5, vjust = -1, face = "bold"), 
          legend.key.size = unit(0.2, "cm"),
          legend.title.position = 'top') + 
    geom_fruit(
        data = ring,
        geom = geom_bar,
        mapping = aes(y = label, x = fitness, fill = fitness),
        orientation = "y",
        stat = "identity",
        width = 0.3
      ) +
    scale_fill_gradient(low = "#93C477", high = "#1F78B4") + 
    guides(fill = guide_colorbar(order = 1,
                                 barwidth = unit(6, "cm")))
  
  
  d <- p$data
  tip_order <- rev(order(d$y[1:length(tree$tip.label)]))
  

  mp_ht <- Heatmap(
    mp[tip_order, ],
    cluster_rows = FALSE,
    cluster_columns = FALSE,
    col = c("#deebf7","#3182bd"),
    column_names_rot = 90,
    border = TRUE
  )


#svg(paste0("Andrew2025/Results/Figures/Truncal_Fitness/OV-",OV_num,"-mp.svg"), width = 4, height = 6)

mp_ht
#dev.off()
  
  #ggsave(paste0("Andrew2025/Results/Figures/OV-",OV_num,"_Fitness_c.svg"),plot = p, width = 6, height = 8)
  
  
  
results_file_u <- paste0(dir_name,"pFitness_ultra.rds")
Results_u <- readRDS(results_file_u)
tree_u <- Results_u$phylo
meanFitness_u <- Results_u$meanFitness[1:length(tree$tip.label)]
mp_u <- Results_u$marginalProb[1:length(tree$tip.label),]

km_HL <- kmeans(meanFitness_u, centers = Num_cluster[i], nstart = 10)


order_idx <- order(km_HL$centers[,1])
new_cluster <- match(km_HL$cluster, order_idx)
km_HL$cluster <- new_cluster
km_HL$centers <- km_HL$centers[order_idx,]

#H_group = 4


types <- ifelse(km_HL$cluster == Num_cluster[i], "High", "Low")
tree_u$ftypes <- types

# if (i == 4){
#   idx_1 <- which(meanFitness_u > 1.96)
#   idx_2 <- which(meanFitness_u > 1.78 & meanFitness_u < 1.85 & seq_along(meanFitness_u) > 200 )
#   
#   types <- rep("Low", times = length(meanFitness_u))
#   types[idx_1] <- "High"
#   types[idx_2] <- "High"
# }else if(i == 3){
#   idx_1 <- which(meanFitness_u > 2.2)
#   idx_2 <- which(meanFitness_u > 2.0 & meanFitness_u < 2.2 & seq_along(meanFitness_u) > 300 )
#   
#   types <- rep("Low", times = length(meanFitness_u))
#   types[idx_1] <- "High"
#   types[idx_2] <- "High"
# }

tip_group_list <- split(tree_u$tip.label,types)


grouped_tree <- groupOTU(tree_u, tip_group_list)
  
ring_u <- tibble(
  label = tree_u$tip.label,
  fitness = meanFitness_u
)
  



  
  p_u <- ggtree(grouped_tree, aes(color = group), size = 0.3)+ ggtitle(paste0("OV-",OV_num," Fitness: (\u03C1: ", format(round(cor(meanFitness,meanFitness_u,method = 'spearman'), 4), nsmall = 4), ")")) +
    scale_color_manual(values = c("High" = "#A13F55","Low" = "black")) +
    labs(fill = "Fitness") +
    theme(legend.position = "bottom", 
          plot.title = element_text(size = 12, hjust = 0.5, vjust = -1, face = "bold"), 
          legend.key.size = unit(0.2, "cm"),
          legend.title.position = 'top') + 
    geom_fruit(
        data = ring_u,
        geom = geom_bar,
        mapping = aes(y = label, x = fitness, fill = fitness),
        orientation = "y",
        stat = "identity",
        width = 0.3
      ) +
    scale_fill_gradient(low = "#93C477", high = "#1F78B4") + 
    guides(fill = guide_colorbar(order = 1,
                                 barwidth = unit(6, "cm")))
  
  #ggsave(paste0("Andrew2025/Results/Figures/OV-",OV_num,"_Fitness_uc.svg"),plot = p_u, width = 6, height = 8)
  
  d_u <- p_u$data
  tip_order_u <- rev(order(d_u$y[1:length(tree_u$tip.label)]))
  
  mp_u_ht <- Heatmap(
    mp_u[tip_order_u,],
    cluster_rows = FALSE,
    cluster_columns = FALSE,
    col = c("#deebf7","#3182bd"),
    column_names_rot = 90,
    border = TRUE
  )
    
  
  #svg(paste0("Andrew2025/Results/Figures/Truncal_Fitness/OV-",OV_num,"-mp_u.svg"), width = 4, height = 6)
  
  mp_u_ht
  #dev.off()
  

  p_output <- p|p_u
  
  
  #ggsave(paste0("Andrew2025/Results/Figures/Truncal_Fitness/OV-",OV_num,"_Fitness_rec.svg"),plot = p_output, width = 12, height = 8)
  

```





## Figure 6(Patient aneuploidy pattern)

To reproduce Figure 6, one need to access the processed copy number state data from the original study [@mcpherson_ongoing_2025] (accession number: syn66366960). 
Specifically, one need to allocate the `signals_SPECTRUM-OV-xxx.h5` file downloaded from Synapse in the directory `Results/patient results/Truncal/OV-xxx`, where `xxx` is the patient id.
This file will be loaded as `sig_data` for each Truncal WGD patient we studied in the following coding block.
After loading the CN state data, one can reproduce the Figure 6 panel A and panel B in the manuscript using the corresponding coding block.

### Load the copy number information


```{r}

Num_cluster <- c(3,2,3,3,2,3,2,3,2,4)#c(2,2,2,2,2,2,2,2,2,2)#c(3,2,3,3,2,3,2,3,2,4)
OV_case <- c("003","008","014", "049","068","082","087","105","118","129")
case_idx <- 9

OV_num <- OV_case[case_idx]

  sig_file  <- paste0("/projects/standard/ruping/shared/wu000766/LeafRank/Andrew2025/signals_SPECTRUM-OV-",OV_num,".h5")
  sig_ls    <- h5ls(sig_file)
  sig_data <- h5read(sig_file,'/')
  
  dir_name <- paste0("Results/patient results/Truncal/OV-",OV_num,"/")
  results_file <- paste0(dir_name,"pFitness_ultra.rds")
  
  Results <- readRDS(results_file)
  
  tree <- Results$phylo
  meanFitness <- Results$meanFitness[1:length(tree$tip.label)]
  mp <- Results$marginalProb[1:length(tree$tip.label),]
  hc_tree <- as.hclust(tree)
  p_fit_ht <- Heatmap(mp,cluster_rows = hc_tree, cluster_columns = FALSE)
  #pdf(paste0(dir_name,"mp_plot.pdf"), width = 13, height = 8)
  #draw(p_fit_ht)
  #dev.off()
  
  tree_labels  <- tree$tip.label
  cell_labels  <- sig_data$obs$brief_cell_id
  n_wgd <- sig_data$obs$n_wgd
  dict_wgd <- as.list(n_wgd)
  names(dict_wgd) <- cell_labels
  sample_wgd <- dict_wgd[tree_labels]
  sampled_idx  <- which(cell_labels %in% tree_labels)
  cell_copys   <- sig_data$layers$copy
  cell_state   <- sig_data$layers$state
  total_count  <- sig_data$layers$totalcounts
  s_cell_copys <- cell_copys[,sampled_idx]
  s_cell_state <- cell_state[,sampled_idx]
  
  
  bin_num      <- sig_data$var$bin
  bin_chr      <- sig_data$var$chr$categories[sig_data$var$chr$codes+1]
  
  
  
  s_cell_state <- t(s_cell_state)
  colnames(s_cell_state) <- bin_num
  col_anno <- data.frame(Chromosome = bin_chr)
  rownames(col_anno) <- colnames(s_cell_state)
  rownames(s_cell_state) <- cell_labels[sampled_idx]
  s_cell_state <- s_cell_state[tree$tip.label,]
  
  n_bin_chr <- bin_chr
  n_bin_chr[n_bin_chr == "X"] <- "23"
  n_bin_chr <- as.numeric(n_bin_chr)
  bin_num_sorted <- bin_num[order(n_bin_chr)]
  n_bin_chr_sorted <- n_bin_chr[order(n_bin_chr)]
  s_cell_state <- s_cell_state[,bin_num_sorted]
  
  col_fun <- colorRamp2(
    c(0,1,2,4,6,8,12),
    c("darkblue","skyblue", "white", "pink", "red","darkred","purple")
  )
  
  #saveRDS(s_cell_state, paste0(dir_name,"copy_numbers.rds"))
```

### Figure 6 Panel A

Panel A is plotted as "ht", which record the heatmap alone with the discrete grouping "mf_K_groups" according to kmeans.
In parallel, we also have "ht_cont" as continuous fitness plot alone with the heatmap.

```{r}


mf_K_groups <- kmeans(meanFitness, centers = Num_cluster[case_idx], nstart = 10)


order_idx <- order(mf_K_groups$centers[,1])
new_cluster <- match(mf_K_groups$cluster, order_idx)
mf_K_groups$cluster <- new_cluster
mf_K_groups$centers <- mf_K_groups$centers[order_idx,]

centers <- mf_K_groups$centers

cluster_labels <- paste0(
  1:length(centers), 
  ": ", 
  round(centers, 2)
)
cluster_labels


mf_K_groups$cluster <- cluster_labels[mf_K_groups$cluster]
  

hc_tree <- as.hclust(tree)

group_colors <- setNames(colorRampPalette(brewer.pal(3, "Set2"))(length(mf_K_groups$centers)), unique(mf_K_groups$cluster))



ha_row <- rowAnnotation(
  Group = mf_K_groups$cluster,
  col = list(Group = group_colors),
  annotation_name_side = "top"
)



ht <- Heatmap(
  s_cell_state,
  name = "Copy Number",
  col = col_fun,               
  cluster_columns = FALSE,
  cluster_rows = hc_tree,
  left_annotation = ha_row,    
  show_row_names = FALSE,
  show_column_names = FALSE,
  column_split = n_bin_chr_sorted,
  border = TRUE
)

#color for fitness grouping
  group_colors <- setNames(colorRampPalette(c("#a8ddb5","#43a2ca"))(length(mf_K_groups$centers)), unique(mf_K_groups$cluster)) # Low #a8ddb5 High #43a2ca

  ha_row <- rowAnnotation(
    Group = mf_K_groups$cluster,
    col = list(Group = group_colors),
    annotation_name_side = "top"
  )

  #color for copy number states
  cn_colors_sun <- c(
    "#000000", # 0: Total Loss (Black)
    "#3182bd", # 1: Heterozygous Loss (Steel Blue)
    "#f0f0f0", # 2: Diploid (Neutral Grey)
    brewer.pal(9, "RdPu") # 3-11: Gains 
    )
    cn_colors_sun <- c(cn_colors_sun, "#67000d")
    col_fun <- colorRamp2(
      0:12,
      cn_colors_sun
    )

  #heatmap
  ht <- Heatmap(
    s_cell_state,
    name = "Copy Number",
    col = col_fun,               
    cluster_columns = FALSE,
    cluster_rows = hc_tree,
    row_dend_width = unit(5, "cm"),   #what I added
    left_annotation = ha_row,
    show_row_names = FALSE,
    show_column_names = FALSE,
    column_split = n_bin_chr_sorted,
    border = TRUE
  )

fit_colors <-  colorRamp2(c(min(meanFitness), max(meanFitness)), c("skyblue", "darkblue"))

ha_row_cont <- rowAnnotation(
  Fitness = meanFitness,
  col = list(Fitness = fit_colors),
  annotation_name_side = "top"
)

ht_cont <- Heatmap(
  s_cell_state,
  name = "Copy Number",
  col = col_fun,               
  cluster_columns = FALSE,
  cluster_rows = hc_tree,
  left_annotation = ha_row_cont,    
  show_row_names = FALSE,
  show_column_names = FALSE,
  column_split = n_bin_chr_sorted,
  border = TRUE
)

#pdf(paste0(dir_name,"Heat_map.pdf"), width = 13, height = 8)
#draw(ht, column_title = paste0("OV-", OV_num))
#dev.off()


#pdf(paste0(dir_name,"Heat_map_cont.pdf"), width = 13, height = 8)
#draw(ht_cont, column_title = paste0("OV-", OV_num))
#dev.off()

#svg(paste0(dir_name,"Heat_map.svg"), width = 12, height = 8)
#draw(ht, column_title = paste0("OV-", OV_num))
#dev.off()

```

### Figure 6 Panel B

Panel B is plotted as "p_HL".

```{r}
Num_cluster <- c(3,2,3,3,2,3,2,2,2,4)#c(2,2,2,2,2,2,2,2,2,2)#c(3,2,3,3,2,3,2,3,2,4)
OV_case <- c("003","008","014", "049","068","082","087","105","118","129")
case_idx <- 7
OV_num <- OV_case[case_idx]

dir_name <- paste0("Results/patient results/Truncal/OV-",OV_num,"/")
results_file <- paste0(dir_name,"pFitness_ultra.rds")

Results <- readRDS(results_file)

tree <- readRDS(paste0(dir_name,"ultra_tree.rds"))
meanFitness <- Results$meanFitness[1:length(tree$tip.label)]

CN_file <- paste0(dir_name,"copy_numbers.rds")
s_cell_state <- readRDS(CN_file)


# km_HL <- kmeans(meanFitness, centers = Num_cluster[case_idx], nstart = 10)
# 
# 
# order_idx <- order(km_HL$centers[,1])
# new_cluster <- match(km_HL$cluster, order_idx)
# km_HL$cluster <- new_cluster
# km_HL$centers <- km_HL$centers[order_idx,]

#H_group = 4


s_H <- s_cell_state[tree$ftypes == "High", ]
s_L <- s_cell_state[tree$ftypes == "Low", ]


# if (case_idx == 4){
#   idx_1 <- which(meanFitness > 1.96)
#   idx_2 <- which(meanFitness > 1.78 & meanFitness < 1.85 & seq_along(meanFitness) > 200 )
#   
#   s_H <- s_cell_state[c(idx_1,idx_2),]
#   s_L <- s_cell_state[setdiff(1:length(meanFitness), c(idx_1,idx_2)),]
# }else if(case_idx == 3){
#   idx_1 <- which(meanFitness > 2.2)
#   idx_2 <- which(meanFitness > 2.0 & meanFitness < 2.2 & seq_along(meanFitness) > 300 )
#   
#   s_H <- s_cell_state[c(idx_1,idx_2),]
#   s_L <- s_cell_state[setdiff(1:length(meanFitness), c(idx_1,idx_2)),]
# }


HL_df <- rbind(data.frame(s_H,group = "H"),data.frame(s_L, group = "L"))

mean_diff <- c()
pvals <- c()
dir   <- c()
GL <- c()

HL_diff_df <- data.frame(
  bin = colnames(s_cell_state),
  Chromosomes = 1:length(colnames(s_cell_state)),
  chr = n_bin_chr_sorted
)

pot_err <- rep(0,length(colnames(s_cell_state)))

n_high <- sum(HL_df$group == "H")
n_low  <- sum(HL_df$group == "L")

for (i in 1:length(colnames(s_cell_state))){
  mean_diff <- c(mean_diff,mean(HL_df[HL_df$group == "H",i]) - mean(HL_df[HL_df$group == "L",i]))
  WT <- wilcox.test(HL_df[HL_df$group == "H",i], HL_df[HL_df$group == "L",i])
  pvals <- c(pvals, WT$p.value)
  temp_df <- data.frame(
    value = HL_df[,i],
    group = HL_df$group
  )
  
  temp_df$group <- factor(
    temp_df$group,
    levels = c('H', 'L')
  )
  
  if (mean(HL_df[HL_df$group == "H",i]) - mean(HL_df[HL_df$group == "L",i]) > 0){
    GL <- c(GL,"Gain")
  }else{
    GL <- c(GL,"Loss")
  }
}

HL_diff_df$mean <- mean_diff
HL_diff_df$wilcox_p <- p.adjust(pvals, method = "BH")
HL_diff_df$CNV <- GL





OV_gene_hl <- data.frame(
  chr = as.numeric(OV_gene$chr),
  start = OV_gene$start_idx,
  end = OV_gene$end_idx,
  label = OV_gene$name
)


p_HL <- ggplot(HL_diff_df,aes(x = Chromosomes,y = wilcox_p, color = CNV)) + geom_point() + geom_hline(yintercept = 0.001) + scale_y_log10() + facet_wrap(~ chr, scales = "free_x", nrow=1) + theme_minimal() + theme(axis.text.x = element_blank()) + labs(title = paste0("OV-",OV_num,": Fraction of high fitness group: ", format(dim(s_H)[1]/dim(s_cell_state)[1], digits = 3)))  

HL_diff_df$pos_mb <- HL_diff_df$Chromosomes * 0.5

lg_cols <- c("Loss" = "#3182bd", "Gain" = brewer.pal(9, "RdPu")[5])
p_HL <- ggplot(HL_diff_df, aes(x = pos_mb, y = wilcox_p, color = CNV, group = interaction(chr, CNV))) + 
  geom_step(direction = "mid") + 
  geom_hline(yintercept = 0.001, linetype = "dashed", color = "red") + 
  scale_y_log10() + 
  
  # Switch to facet_grid and set space to "free_x"
  facet_grid(. ~ chr, scales = "free_x", space = "free_x") +
  scale_color_manual(values = lg_cols) +
  
  theme_minimal() + 
  theme(
    axis.text.x = element_blank(),
    panel.spacing = unit(0.1, "lines"), # Optional: Reduce space between chromosomes
    strip.background = element_rect(fill = "grey90") # Makes chr labels stand out
  ) + 
  labs(title = paste0("OV-", OV_num, ": Fraction of high fitness group: ", 
                      format(dim(s_H)[1]/dim(s_cell_state)[1], digits = 3)))



#p_HL <- p_HL +
#  geom_vline(data = OV_gene_hl, aes(xintercept = start), color = "orange", linetype = "dashed") +
#  geom_vline(data = OV_gene_hl, aes(xintercept = end), color = "orange", linetype = "dashed") +
#  geom_text(
#    data = OV_gene_hl,
#    aes(x = (start + end)/2, y = Inf, label = label),
#    vjust = 1.5, size = 3, inherit.aes = FALSE
#  )
#ggsave("HL_pval.svg",p_HL,width = 12, height = 5.2)

#p_HL <- p_HL +
#  geom_vline(data = OV_gene_hl, aes(xintercept = start), color = "orange", linetype = "dashed") +
#  geom_vline(data = OV_gene_hl, aes(xintercept = end), color = "orange", linetype = "dashed") +
#  geom_text(
#    data = OV_gene_hl,
#    aes(x = (start + end)/2, y = Inf, label = label),
#    vjust = 1.5, size = 3, inherit.aes = FALSE
#  )
#ggsave(paste0(dir_name,"HL_pval_final.svg"),p_HL,width = 12, height = 6)

#ggsave(paste0(dir_name,"mean_diff.pdf"),p_mean_diff,width = 8, height = 6)

#saveRDS(HL_diff_df,paste0(dir_name,"HL_pval.rds"))






```



### Figure 6 Panel B cross patient PTEN focal region

Panel B cross patient PTEN focal region is plotted as "p_IG".
The "IG" can be modified to examine other gene across the patient.

```{r}



OV_Truncal <- c("003","008", "014", "049", "068", "082","087", "105", "118", "129")
Num_cluster <- c(3,2,3,3,2,3,2,2,2,4)#c(3,2,3,3,2,3,2,3,2,4)

#Chr_idx <- c("1","2","3","4","5","6","7","8","9","10","11","12","13","14","15","16","17","18","19","20","21","22","X")

#for (chr_i in Chr_idx){

#IG <- paste0("Chr",chr_i)


#idx = which(startsWith(colnames(s_cell_state), paste0(chr_i,":")))

# The IG_start and IG_end are obtained from the consecutive gain test obtained from the next section. One can manually modified this to examine other regions.

IG <- "PTEN"
IG_loc_hg38 <- c(87863625, 87971930)
IG_loc_hg19 <- c(89623382, 89731687)
IG_start <- which(colnames(s_cell_state) == "10:38000001-38500000")
IG_end   <- which(colnames(s_cell_state) == "10:99500001-100000000")
#
xx_start <- 1 + (IG_end - IG_start-1)*((IG_loc_hg19[1]-38250000) / (99750000 - 38250000))
xx_end <- 1 + (IG_end - IG_start-1)*((IG_loc_hg19[2]-38250000) / (99750000 - 38250000))

# 
# IG <- "KRAS"
# IG_loc_hg38 <- c(25205246, 25250929)
# IG_loc_hg19 <- c(25353612, 25408431)
# IG_start <- which(colnames(s_cell_state) == "12:6500001-7000000")
# IG_end   <- which(colnames(s_cell_state) == "12:27500001-28000000")
# 
# xx_start <- 1 + (IG_end - IG_start-1)*((IG_loc_hg19[1]-6500001) / (27750000 - 6500001))
# xx_end <- 1 + (IG_end - IG_start-1)*((IG_loc_hg19[2]-6500001) / (27750000 - 6500001))


# IG <- "MAP3K7"
# IG_loc_hg38 <- c(90513579, 90587072)
# IG_loc_hg19 <- c(91215949, 91304140)
# IG_start <- which(colnames(s_cell_state) == "6:88000001-88500000")
# IG_end   <- which(colnames(s_cell_state) == "6:103500001-104000000")
# 
# xx_start <- 1 + (IG_end - IG_start-1)*((90513579-88250000) / (103750000 - 88250000))
# xx_end <- 1 + (IG_end - IG_start-1)*((90587072-88250000) / (103750000 - 88250000))


#IG_start <- min(idx)
#IG_end   <- max(idx)




df_IG <- data.frame()

y_pos <- 0
y_tick_breaks <- c()
y_r_tick_breaks <- c()
y_tick_labels <- c()
y_r_tick_labels <- c()

temp <- c()
for (ii in 1:length(OV_Truncal)){
  OV_num <- OV_Truncal[ii]
  cluster_num <- Num_cluster[ii]

  
  
  dir_name <- paste0("Results/patient results/Truncal/OV-",OV_num,"/")
  results_file <- paste0(dir_name,"pFitness_ultra.rds")
  CN_file <- paste0(dir_name,"copy_numbers.rds")
  CN_p_file <- paste0(dir_name,"HL_manu.rds")
  # if (ii %in% alt_list){
  #   CN_p_file <- paste0(dir_name,"HL_final_alt.rds")
  # }else if (ii == 11){
  #   CN_p_file <- paste0(dir_name,"LH_final.rds")
  # }else if (ii %in% alt_list_2){
  #   CN_p_file <- paste0(dir_name,"HL_final_alt2.rds")
  # }
    
  Results <- readRDS(results_file)
    
  tree <- readRDS(paste0(dir_name,"ultra_tree.rds"))
  meanFitness <- Results$meanFitness[1:length(tree$tip.label)]
  
  
  
  # mf_K <- kmeans(meanFitness,centers = cluster_num, nstart = 10)
  # order_idx <- order(mf_K$centers[,1])
  # new_cluster <- match(mf_K$cluster, order_idx)
  # mf_K$cluster <- new_cluster
  # mf_K$centers <- mf_K$centers[order_idx,]
  
  CN_mat <- readRDS(CN_file)
  CN_p_df <- readRDS(CN_p_file)
  CN_p_df <- CN_p_df[IG_start:IG_end,]
  
  CN_H_IG <- CN_mat[tree$ftype == "High", IG_start:IG_end]
  CN_L_IG <- CN_mat[tree$ftype == "Low", IG_start:IG_end]
  
  # if (OV_num == "041"){
  #   CN_H_IG <- CN_mat[mf_K$cluster >=2, IG_start:IG_end]
  #   CN_L_IG <- CN_mat[mf_K$cluster ==1, IG_start:IG_end]
  # }else{
  #   CN_H_IG <- CN_mat[mf_K$cluster == cluster_num,IG_start:IG_end]
  #   CN_L_IG <- CN_mat[mf_K$cluster != cluster_num,IG_start:IG_end]
  # }
  # if (OV_num == "014" & ii %in% alt_list_2){
  #   idx_1 <- which(meanFitness > 2.2)
  #   idx_2 <- which(meanFitness > 2.0 & meanFitness < 2.2 & seq_along(meanFitness) > 300 )
  #   
  #   CN_H_IG <- CN_mat[c(idx_1,idx_2),IG_start:IG_end]
  #   CN_L_IG <- CN_mat[setdiff(1:length(meanFitness), c(idx_1,idx_2)),IG_start:IG_end]
  # }else if (OV_num == "049" & ii %in% alt_list_2){
  #   idx_1 <- which(meanFitness > 1.96)
  #   idx_2 <- which(meanFitness > 1.78 & meanFitness < 1.85 & seq_along(meanFitness) > 200 )
  #   
  #   CN_H_IG <- CN_mat[c(idx_1,idx_2),IG_start:IG_end]
  #   CN_L_IG <- CN_mat[setdiff(1:length(meanFitness), c(idx_1,idx_2)),IG_start:IG_end]
  # }
  
  
  CN_H_IG_mean <- apply(CN_H_IG,2,mean)
  CN_L_IG_mean <- apply(CN_L_IG,2,mean)
  
  df_IG_sub <- data.frame(
    patient_idx = rep(ii,IG_end-IG_start+1),
    IG = 0:(IG_end-IG_start),
    deltaCN_y = (CN_H_IG_mean - CN_L_IG_mean) + y_pos,
    deltaCN = (CN_H_IG_mean - CN_L_IG_mean),
    LowCN   = CN_L_IG_mean,
    HighCN  = CN_H_IG_mean
  )
  
  
  Significance <- c()
  
  for (i in 1:length(colnames(CN_H_IG))){
    pvals <-  CN_p_df$wilcox_p[i]
    if (pvals >=0.05 || is.nan(pvals)){
      Significance <- c(Significance,"ns")
    } else if (pvals>=0.01){
      Significance <- c(Significance, "*")
    } else if (pvals>=0.001){
      Significance <- c(Significance, "**")
    } else if (pvals>=0.0001){
      Significance <- c(Significance, "***")
    } else {
      Significance <- c(Significance, "****")
    }
  }
  
  df_IG_sub$Significance <- Significance
  
  df_IG <- rbind(df_IG,df_IG_sub)
  
  y_tick_breaks <- c(y_tick_breaks,c(y_pos-1, y_pos, y_pos+1))
  y_r_tick_breaks <- c(y_r_tick_breaks,y_pos)
  y_tick_labels <- c(y_tick_labels,c("\u2212 1",paste0(format(mean(CN_L_IG_mean),digits = 2)),"+ 1"))
  y_r_tick_labels <- c(y_r_tick_labels,paste0(OV_num,"(",format(dim(CN_H_IG)[1]/dim(CN_mat)[1],digits = 2),")"))
  
  y_pos <- y_pos + 4

}


#IG_check <- c()

#for (i in 0:length(idx)-1){
#  check_i   <- df_IG[df_IG$IG == i,]
#  check_sig <- check_i[check_i$Significance == #"****"|check_i$Significance == "***"|check_i$Significance == "**", ]
#  if (dim(check_sig)[1] >0){
#    if (all(check_sig$deltaCN > 0)| all(check_sig$deltaCN <0)){
#      IG_check <- c(IG_check,idx[i+1])
#    }
#  }
#}

#IG_idx <- c(IG_idx,IG_check)



stat_color <- c(
  "ns"    = "#BDBDBD",
  "*"     = "#A6CEE3",
  "**"    = "#1F78B4",
  "***"   = "#E31A1C",
  "****"  = "#B10026"
)



p_IG <- ggplot(df_IG,aes(x = IG, y = deltaCN_y, size = Significance)) + geom_hline(yintercept = y_r_tick_breaks, color = "#a8ddb5", linewidth = 2) + geom_point(color = "#43a2ca") + scale_size_manual(values = c("ns" = 0.5, "*" = 3, "**" = 4, "***" = 5, "****" = 6))    + scale_y_continuous(
  name   = "CN",
  breaks = y_tick_breaks,
  labels = y_tick_labels,
  limits = c(min(y_tick_breaks), max(y_tick_breaks) ),
  sec.axis = sec_axis(
    trans  = ~.,
    breaks = y_r_tick_breaks,
    name   = "Patient #",
    labels = y_r_tick_labels
  ) 
) + 
  geom_vline(
    xintercept = xx_end + (xx_start - xx_end)/2,
    linetype = "dashed",
    linewidth = 0.6
  ) +
  scale_x_continuous(name = IG)  + theme(axis.text.y = element_text(face = "bold", family = "mono", size = 15), axis.text.x = element_blank(), axis.ticks.x = element_blank(), text = element_text(face = "bold"), axis.text.y.right = element_text(angle = 270, vjust = 0.5, hjust = 0.5)) 








```







### Supplementary Figures




#### (Figure 5 panel C and D)Plotting the PCA clustering

To reproduce the figure 5 panel C and D, one can use the following code section with loaded fitness prediction and CN state to re-do the PCA analysis.


```{r}

  if (case_idx == 9){
    mf_K_groups <- kmeans(meanFitness, centers = 4  , nstart = 50)
  }else{
    mf_K_groups <- kmeans(meanFitness, centers = Num_cluster[case_idx]  , nstart = 50)
  }
    
  
  
  
  order_idx <- order(mf_K_groups$centers[,1])
  new_cluster <- match(mf_K_groups$cluster, order_idx)
  mf_K_groups$cluster <- new_cluster
  mf_K_groups$centers <- mf_K_groups$centers[order_idx,]
  
  centers <- mf_K_groups$centers
  
  cluster_labels <- paste0(
    1:length(centers), 
    ": ", 
    round(centers, 2)
  )
  cluster_labels
  
  
  mf_K_groups$cluster <- cluster_labels[mf_K_groups$cluster]
  
  mf_ring <- tibble(
    label = tree$tip.label,
    fitness = meanFitness
  )
  

  
  
  
  
  s_cell_state_pca <- s_cell_state[, apply(s_cell_state, 2, var) !=0]
  pca <- prcomp(s_cell_state_pca, center=TRUE, scale.=TRUE)
  
  
  
  cluster_color <- c("#DDC48E", "#7E4F59", "#A13F55", "#333333")
  
  
  
  pca_2d <-data.frame(pca$x[,1:2])
  colnames(pca_2d) <- c("PC1", "PC2")
  pca_mc <- Mclust(pca_2d)
  pca_2d$cluster <- factor(pca_mc$classification)
  centers <- pca_2d %>%
    group_by(cluster) %>%
    summarise(
      PC1 = mean(PC1),
      PC2 = mean(PC2)
    )
  hulls<- pca_2d %>%
    group_by(cluster) %>%
    slice(chull(PC1,PC2))
  
  g_list <- c()
  for (i in 1:length(mf_K_groups$cluster)){
    g_list <- c(g_list,paste0(as.character(mf_K_groups$cluster[i])))
  }
  
  pca_2d$Group <- g_list
  
  p_pca <- ggplot(pca_2d, aes(x = PC1, y = PC2, color = Group)) +
    geom_point(size = 2) +
    scale_color_manual(values = cluster_color) +
    #stat_ellipse(aes(group = cluster), linewidth = 1) + 
    #geom_point(data = centers, aes(PC1,PC2), shape= 4, size = 5, stroke=1.5, color = "black") +
    theme_classic() + 
    labs(title = paste0("OV-", OV_num,":PCA projection colored by group"))
  



```



#### (Figure s5 panel A) Overall aneuploidy analysis across patient

In this section, we analyze the overall aneuploidy pattern across the Truncal WGD patients.
We first examine the average copy number levels at each bins.

```{r}

OV_Truncal <- c("003","008", "014", "049", "068", "082","087", "105", "118", "129")
#OV_Truncal <- c("118")
Num_cluster <- c(1,1,1,1,1,1,1,1,1,1)#c(3,2,3,3,3,4,4,3,4,4)
Chr_idx <- c("1","2","3","4","5","6","7","8","9","10","11","12","13","14","15","16","17","18","19","20","21","22","X")
#Chr_idx <- c("16")

df_IG <- data.frame()

for (chr_i in Chr_idx){
IG <- paste0("Chr",chr_i)


idx = which(startsWith(colnames(s_cell_state), paste0(chr_i,":")))

#IG <- "NF1"
#IG_start <- OV_gene$start_idx[OV_gene$name == IG]
#IG_end   <- OV_gene$end_idx[OV_gene$name == IG]
#IG_start <- which(colnames(s_cell_state) == "4:1-500000")
#IG_end   <- which(colnames(s_cell_state) == "4:49500001-50000000")


IG_start <- min(idx)
IG_end   <- max(idx)




df_IG_sub <- data.frame(
  chr   = rep(chr_i,length(idx)),   
  gain_3  = rep(0,length(idx)),  # CN bins >=7
  gain_2   = rep(0,length(idx)), # CN bins ~ 6
  gain_1   = rep(0,length(idx)), # CN bins ~ 5
  loss_1 = rep(0,length(idx)),   # CN bins ~ 3
  loss_2  = rep(0,length(idx)),  # CN bins ~ 2
  loss_3   = rep(0,length(idx))  # CN bins <= 1
)

rownames(df_IG_sub) <- colnames(s_cell_state)[idx]


  for (ii in 1:length(OV_Truncal)){
    OV_num <- OV_Truncal[ii]
  
    
    
    dir_name <- paste0("Results/patient results/Truncal/OV-",OV_num,"/")
    CN_file <- paste0(dir_name,"copy_numbers.rds")
    CN_p_file <- paste0(dir_name,"HL_manu.rds")
    
    
    CN_mat <- readRDS(CN_file)
    CN_mat_sub <- CN_mat[,IG_start:IG_end]
    
  
    
  
    
    
    
    for (i in 1:length(colnames(CN_mat_sub))){
      CN_mean <- mean(CN_mat_sub[,i])
      if (CN_mean >= 7){
        df_IG_sub$gain_1[i] <- df_IG_sub$gain_1[i] + 1
      } else if (CN_mean >= 6){
        df_IG_sub$gain_1[i] <- df_IG_sub$gain_1[i] + 1
      } else if (CN_mean >= 4.5){
        df_IG_sub$gain_1[i] <- df_IG_sub$gain_1[i] + 1
      } else if (CN_mean <= 1.5){
        df_IG_sub$loss_1[i] <- df_IG_sub$loss_1[i] + 1
      } else if (CN_mean <= 2.5){
        df_IG_sub$loss_1[i] <- df_IG_sub$loss_1[i] + 1
      } else if (CN_mean <= 3.5){
        df_IG_sub$loss_1[i] <- df_IG_sub$loss_1[i] + 1
      }
    }
    
    
    
  }



df_IG <- rbind(df_IG, df_IG_sub)


}



```

The aneuploidy across 10 truncal patient are plot here as p1


```{r}

df_IG$seg <- rownames(df_IG)



df <- data.frame(
  chrom = df_IG$chr,
  segment = df_IG$seg,
  gain_count2 = df_IG$gain_1,
  loss_count2 = df_IG$loss_1
)

df2 <- df %>%
  mutate(
    chrom = as.character(chrom),
    seg_start = as.numeric(str_extract(segment, "(?<=:)\\d+"))
  ) %>%
  group_by(chrom) %>%
  arrange(seg_start, .by_group = TRUE) %>%              # small → large
  mutate(
    segment_f = factor(segment, levels = unique(segment)) # freeze order
  ) %>%
  ungroup()








seg_runs <- data.frame()

start_seg <- 0
end_seg <- 0

prev_chrom <- 1
chrom_pos  <- 1
for (i in 1:dim(df)[1]){
  if (df[i,1] != prev_chrom){
    if (end_seg >= start_seg+30 & start_seg !=0){
      
      new_rows <- data.frame(chrom = prev_chrom, xmin = start_seg-chrom_pos, xmax = end_seg-chrom_pos, ymin = -Inf, ymax = Inf)
      seg_runs <- rbind(seg_runs,new_rows)
      
    }
    prev_chrom <- df[i,1]
    chrom_pos <- i
    start_seg <- 0
    end_seg <- 0
  } 
  
  
  
  is_mono <- (sum(df[i,seq(3,ncol(df), by = 2)])>=3 & sum(df[i,seq(4,ncol(df), by = 2)])==0)||(sum(df[i,seq(3,ncol(df), by = 2)])==0 & sum(df[i,seq(4,ncol(df), by = 2)])>=3)
  if (is_mono){
    if (start_seg == 0){
      start_seg <- i
    }else{
      end_seg <- i
    }
  }else{
    if (end_seg >= start_seg+30 & start_seg !=0){
      new_rows <- data.frame(chrom = df[i,1], xmin = start_seg-chrom_pos, xmax = end_seg-chrom_pos, ymin = -Inf, ymax = Inf)
      seg_runs <- rbind(seg_runs,new_rows)
    }
    start_seg <- 0
    end_seg <- 0
  }
}





chrom_levels <- c(as.character(1:22), "X")

df_long <- df2 %>%
  pivot_longer(cols = c(gain_count2,loss_count2),
               names_to = "type",
               values_to = "count") %>%
  mutate(count = ifelse(type == "loss_count2", -count, count))

df_long$chrom <- factor(df_long$chrom, levels = chrom_levels)
df_long$type <- as.factor(df_long$type)
df_long <- df_long %>%
  group_by(chrom) %>%
  mutate(seg_id = dense_rank(segment_f)) %>%
  ungroup()


df_sum <- df_long %>%
  group_by(chrom,seg_id) %>%
  summarise(sum_count = sum(count), .groups = "drop")

seg_runs$chrom <- factor(seg_runs$chrom, levels = chrom_levels)

w <- 0.5
p1<- ggplot(df_long, aes(x = seg_id, y = count, fill = type)) +
  geom_col(width = 1) +
  #geom_line(data = df_sum,
  #          aes(x = seg_id,y = sum_count, group= chrom),
  #          inherit.aes = FALSE,
  #          linewidth = 0.9,
  #          color = "grey20")+
  facet_grid(. ~ chrom, scales = "free_x", space = "free_x") +
  #geom_hline(yintercept = 0, color = "black") +
  scale_fill_manual(
    values = c("gain_count2" = "#B11226",
               "loss_count2" = "#1F4E8C"),
    labels = c("Gain","Loss")
  ) +
  labs(
    x = "Segment",
    y = "Number of Patients",
    title = "Cross-patient high fitness CN pattern"
  ) +
  theme_bw() +
  theme(
    strip.background = element_rect(fill = "lightgrey"),
    panel.spacing = unit(0.15, "lines"),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.title.x = element_blank(),
    axis.title.y = element_text(face = "bold"),
    plot.title = element_text(face = "bold")
  ) #+
  #geom_rect(
  #  data = seg_runs,
  #  aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
  #  inherit.aes = FALSE,
  #  fill = '#F2E6B8',
  #  alpha = 0.5
  #)



#"gain_count4" = "#DE2D26",  
#               "loss_count4" = "#2171B5",
#               "gain_count3" = "#FC9272",
#               "loss_count3" = "#6BAED6",
#               "gain_count2" = "#FEE0D2",
#               "loss_count2" = "#C6DBEF"

#ggsave("Andrew2025/Results/Figures/Cross-patient_plot.pdf", plot = p, width = 15, height = 6)

```

#### (Figure s5 panel B)

The cross-patient aneuploidy pattern data recorded in df_IG, where each row represents one bin and each column represents number of patient in certain significant levels.

```{r}



OV_Truncal <- c("003","008", "014", "049", "068", "082","087", "105", "118", "129")
Num_cluster <- c(3,2,3,3,2,3,2,2,2,4)#c(2,2,2,2,2,2,2,2,2,2)#c(3,2,3,3,2,3,2,3,2,4)

Chr_idx <- c("1","2","3","4","5","6","7","8","9","10","11","12","13","14","15","16","17","18","19","20","21","22","X")


df_IG <- data.frame()
s_cell_state <- readRDS('Results/patient results/Truncal/OV-003/copy_numbers.rds')


for (chr_i in Chr_idx){
IG <- paste0("Chr",chr_i)


idx = which(startsWith(colnames(s_cell_state), paste0(chr_i,":")))

#IG <- "NF1"
#IG_start <- OV_gene$start_idx[OV_gene$name == IG]
#IG_end   <- OV_gene$end_idx[OV_gene$name == IG]
#IG_start <- which(colnames(s_cell_state) == "4:1-500000")
#IG_end   <- which(colnames(s_cell_state) == "4:49500001-50000000")


IG_start <- min(idx)
IG_end   <- max(idx)




df_IG_sub <- data.frame(
  chr   = rep(chr_i,length(idx)),
  gain_p0001 = rep(0,length(idx)),
  gain_p001  = rep(0,length(idx)),
  gain_p01   = rep(0,length(idx)),
  gain_p05   = rep(0,length(idx)),
  gain_ns    = rep(0,length(idx)),
  loss_p0001 = rep(0,length(idx)),
  loss_p001  = rep(0,length(idx)),
  loss_p01   = rep(0,length(idx)),
  loss_p05   = rep(0,length(idx)),
  loss_ns    = rep(0,length(idx))
)



rownames(df_IG_sub) <- colnames(s_cell_state)[idx]


  for (ii in 1:length(OV_Truncal)){
    OV_num <- OV_Truncal[ii]
    cluster_num <- Num_cluster[ii]
  
    
    
    dir_name <- paste0("Results/patient results//Truncal/OV-",OV_num,"/")
    results_file <- paste0(dir_name,"pFitness_ultra.rds")
    CN_file <- paste0(dir_name,"copy_numbers.rds")
    CN_p_file <- paste0(dir_name,"HL_manu.rds")
    #if (ii == 4){
    #  CN_p_file <- paste0(dir_name,"LH_final.rds")
    #}
    # if (ii %in% alt_list){
    #   CN_p_file <- paste0(dir_name,"HL_final_alt.rds")
    # }else if (ii %in% alt_list_2){
    #   CN_p_file <- paste0(dir_name,"HL_final_alt2.rds")
    # }
    CN_mat <- readRDS(CN_file)
    CN_p_df <- readRDS(CN_p_file)
      
    Results <- readRDS(results_file)
      
    tree <- readRDS(paste0(dir_name,"ultra_tree.rds"))
    meanFitness <- Results$meanFitness[1:length(tree$tip.label)]
    
    
    
    # mf_K <- kmeans(meanFitness,centers = cluster_num, nstart = 10)
    # order_idx <- order(mf_K$centers[,1])
    # new_cluster <- match(mf_K$cluster, order_idx)
    # mf_K$cluster <- new_cluster
    # mf_K$centers <- mf_K$centers[order_idx,]
    

    CN_p_df <- CN_p_df[IG_start:IG_end,]
    
    CN_H_IG <- CN_mat[tree$ftypes == "High",IG_start:IG_end]
    CN_L_IG <- CN_mat[tree$ftypes == "Low",IG_start:IG_end]
    
    # if (OV_num == "014"){
    #   idx_1 <- which(meanFitness > 2.2)
    #   idx_2 <- which(meanFitness > 2.0 & meanFitness < 2.2 & seq_along(meanFitness) > 300 )
    #   
    #   CN_H_IG <- CN_mat[c(idx_1,idx_2),IG_start:IG_end]
    #   CN_L_IG <- CN_mat[setdiff(1:length(meanFitness), c(idx_1,idx_2)),IG_start:IG_end]
    # }else if (OV_num == "049"){
    #   idx_1 <- which(meanFitness > 1.96)
    #   idx_2 <- which(meanFitness > 1.78 & meanFitness < 1.85 & seq_along(meanFitness) > 200 )
    #   
    #   CN_H_IG <- CN_mat[c(idx_1,idx_2),IG_start:IG_end]
    #   CN_L_IG <- CN_mat[setdiff(1:length(meanFitness), c(idx_1,idx_2)),IG_start:IG_end]
    # }
    
    CN_H_IG_mean <- apply(CN_H_IG,2,mean)
    CN_L_IG_mean <- apply(CN_L_IG,2,mean)
    
    




    for (i in 1:length(colnames(CN_H_IG))){
      pvals <-  CN_p_df$wilcox_p[i]
      if (pvals >=0.05 || is.nan(pvals)){
        if (CN_H_IG_mean[i] > CN_L_IG_mean[i]){
          df_IG_sub$gain_ns[i] <- df_IG_sub$gain_ns[i] + 1
        }else{
          df_IG_sub$loss_ns[i] <- df_IG_sub$loss_ns[i] + 1
        }
      } else if (pvals>=0.01){
        if (CN_H_IG_mean[i] > CN_L_IG_mean[i]){
          df_IG_sub$gain_p05[i] <- df_IG_sub$gain_p05[i] + 1
        }else{
          df_IG_sub$loss_p05[i] <- df_IG_sub$loss_p05[i] + 1
        }
      } else if (pvals>=0.001){
        if (CN_H_IG_mean[i] > CN_L_IG_mean[i]){
          df_IG_sub$gain_p01[i] <- df_IG_sub$gain_p01[i] + 1
        }else{
          df_IG_sub$loss_p01[i] <- df_IG_sub$loss_p01[i] + 1
        }
      } else if (pvals>=0.0001){
        if (CN_H_IG_mean[i] > CN_L_IG_mean[i]){
          df_IG_sub$gain_p001[i] <- df_IG_sub$gain_p001[i] + 1
        }else{
          df_IG_sub$loss_p001[i] <- df_IG_sub$loss_p001[i] + 1
        }
      } else {
        if (CN_H_IG_mean[i] > CN_L_IG_mean[i]){
          df_IG_sub$gain_p0001[i] <- df_IG_sub$gain_p0001[i] + 1
        }else{
          df_IG_sub$loss_p0001[i] <- df_IG_sub$loss_p0001[i] + 1
        }
      }
    }
    
    # for (i in 1:length(colnames(CN_H_IG))){
    #   pvals <-  CN_p_df$wilcox_p[i]
    #   if (pvals >=0.05 || is.nan(pvals)){
    #     if (CN_p_df$dir[i]>=0){
    #       df_IG_sub$gain_ns[i] <- df_IG_sub$gain_ns[i] + 1
    #     }else{
    #       df_IG_sub$loss_ns[i] <- df_IG_sub$loss_ns[i] + 1
    #     }
    #   } else if (pvals>=0.01){
    #     if (CN_p_df$dir[i]>=0){
    #       df_IG_sub$gain_p05[i] <- df_IG_sub$gain_p05[i] + 1
    #     }else{
    #       df_IG_sub$loss_p05[i] <- df_IG_sub$loss_p05[i] + 1
    #     }
    #   } else if (pvals>=0.001){
    #     if (CN_p_df$dir[i]>=0){
    #       df_IG_sub$gain_p01[i] <- df_IG_sub$gain_p01[i] + 1
    #     }else{
    #       df_IG_sub$loss_p01[i] <- df_IG_sub$loss_p01[i] + 1
    #     }
    #   } else if (pvals>=0.0001){
    #     if (CN_p_df$dir[i]>=0){
    #       df_IG_sub$gain_p001[i] <- df_IG_sub$gain_p001[i] + 1
    #     }else{
    #       df_IG_sub$loss_p001[i] <- df_IG_sub$loss_p001[i] + 1
    #     }
    #   } else {
    #     if (CN_p_df$dir[i]>=0){
    #       df_IG_sub$gain_p0001[i] <- df_IG_sub$gain_p0001[i] + 1
    #     }else{
    #       df_IG_sub$loss_p0001[i] <- df_IG_sub$loss_p0001[i] + 1
    #     }
    #   }
    # }
    
    
    
  }



df_IG <- rbind(df_IG, df_IG_sub)


}


df_IG$seg <- rownames(df_IG)



df <- data.frame(
  chrom = df_IG$chr,
  segment = df_IG$seg,
  gain_count4 = df_IG$gain_p0001,
  loss_count4 = df_IG$loss_p0001,
  gain_count3 = df_IG$gain_p001 ,
  loss_count3 = df_IG$loss_p001 ,
  gain_count2 = df_IG$gain_p01 ,
  loss_count2 = df_IG$loss_p01 
)

df2 <- df %>%
  mutate(
    chrom = as.character(chrom),
    seg_start = as.numeric(str_extract(segment, "(?<=:)\\d+"))
  ) %>%
  group_by(chrom) %>%
  arrange(seg_start, .by_group = TRUE) %>%              # small → large
  mutate(
    segment_f = factor(segment, levels = unique(segment)) # freeze order
  ) %>%
  ungroup()








seg_runs <- data.frame()

start_seg <- 0
end_seg <- 0

prev_chrom <- 1
chrom_pos  <- 1
for (i in 1:dim(df)[1]){
  if (df[i,1] != prev_chrom){
    if (end_seg >= start_seg+30 & start_seg !=0){
      
      new_rows <- data.frame(chrom = prev_chrom, xmin = start_seg-chrom_pos, xmax = end_seg-chrom_pos, ymin = -Inf, ymax = Inf, start_name = df$segment[start_seg], end_name = df$segment[end_seg])
      seg_runs <- rbind(seg_runs,new_rows)
      
    }
    prev_chrom <- df[i,1]
    chrom_pos <- i
    start_seg <- 0
    end_seg <- 0
  } 
  
  
  
  is_mono <- (sum(df[i,seq(3,ncol(df), by = 2)])>=3 & sum(df[i,seq(4,ncol(df), by = 2)])==0)||(sum(df[i,seq(3,ncol(df), by = 2)])==0 & sum(df[i,seq(4,ncol(df), by = 2)])>=3)
  if (is_mono){
    if (start_seg == 0){
      start_seg <- i
    }else{
      end_seg <- i
    }
  }else{
    if (end_seg >= start_seg+30 & start_seg !=0){
      new_rows <- data.frame(chrom = df[i,1], xmin = start_seg-chrom_pos, xmax = end_seg-chrom_pos, ymin = -Inf, ymax = Inf, start_name = df$segment[start_seg], end_name = df$segment[end_seg])
      seg_runs <- rbind(seg_runs,new_rows)
    }
    start_seg <- 0
    end_seg <- 0
  }
}





chrom_levels <- c(as.character(1:22), "X")

df_long <- df2 %>%
  pivot_longer(cols = c(gain_count4, loss_count4,gain_count3, loss_count3, gain_count2,loss_count2),
               names_to = "type",
               values_to = "count") %>%
  mutate(count = ifelse(type == "loss_count2" |type == "loss_count3"|type == "loss_count4", -count, count))

df_long$chrom <- factor(df_long$chrom, levels = chrom_levels)
df_long$type <- as.factor(df_long$type)
df_long <- df_long %>%
  group_by(chrom) %>%
  mutate(seg_id = dense_rank(segment_f)) %>%
  ungroup()


df_sum <- df_long %>%
  group_by(chrom,seg_id) %>%
  summarise(sum_count = sum(count), .groups = "drop")

seg_runs$chrom <- factor(seg_runs$chrom, levels = chrom_levels)

w <- 0.5
p<- ggplot(df_long, aes(x = seg_id, y = count, fill = type)) +
  geom_col(width = 1) +
  #geom_line(data = df_sum,
  #          aes(x = seg_id,y = sum_count, group= chrom),
  #          inherit.aes = FALSE,
  #          linewidth = 0.9,
  #          color = "grey20")+
  facet_grid(. ~ chrom, scales = "free_x", space = "free_x") +
  geom_hline(yintercept = 0, color = "black") +
  scale_fill_manual(
    values = c("gain_count4" = "#B11226",  
               "loss_count4" = "#1F4E8C",
               "gain_count3" = "#E34A4A",
               "loss_count3" = "#4F8CC9",
               "gain_count2" = "#F4A6A6",
               "loss_count2" = "#A6C8E5"),
    labels = c("Gain**", "Gain***","Gain****", "Loss**","Loss***", "Loss****")
  ) +
  labs(
    x = "Segment",
    y = "Number of Patients",
    title = "Cross-patient high fitness CN pattern"
  ) +
  theme_bw() +
  theme(
    strip.background = element_rect(fill = "lightgrey"),
    panel.spacing = unit(0.15, "lines"),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.title.x = element_blank(),
    axis.title.y = element_text(face = "bold"),
    plot.title = element_text(face = "bold")
  ) +
  geom_rect(
    data = seg_runs,
    aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
    inherit.aes = FALSE,
    fill = '#F2E6B8',
    alpha = 0.5
  )




```





## Figure 7

### Panel Fitness

The following code section reproduce the fitness inference in Figure 7. The fitness inference is plotted as `p_output`. The heatmap for inferred fitness distribution are plotted as `mp_ht`.

```{r}
OV_num <- '075'

dir_name <- paste0("Results/patient results/Subclonal/OV-", OV_num,"/")




Results_ultra <- readRDS(paste0(dir_name, "pFitness_ultra.rds"))
tree_ultra <- Results_ultra$phylo
mf_ultra <- Results_ultra$meanFitness[1:length(tree_ultra$tip.label)]
mp <- Results_ultra$marginalProb[1:length(tree_ultra$tip.label),]
b_rates <- 1.1*1.1^(0:15)
colnames(mp) <- c(as.character(round(b_rates - 1,2)))
u_tree <- readRDS(paste0(dir_name,"ultra_tree.rds"))
hc_tree <- as.hclust(u_tree)
mp_ht <- Heatmap(
  mp,
  cluster_rows = hc_tree,
  cluster_columns = FALSE,
  col = c("#deebf7","#3182bd"),
  column_names_rot = 90,
  border = TRUE
)

tree_ultra$edge_wgd <- ifelse(tree_ultra$edge_wgd == 1, "WGD +", "WGD -")
edge_groups <- data.frame(
  node = tree_ultra$edge[,2],
  group = tree_ultra$edge_wgd
)
#svg(paste0("Andrew2025/Results/Figures/Subclonal_Fitness/OV-",OV_num,"-mp.svg"), width = 4, height = 6)

#mp_ht
#dev.off()

mf_ring <- tibble(
  label = tree_ultra$tip.label,
  fitness = mf_ultra
)




p_ultra <- ggtree(tree_ultra, layout = "circular", size = 0) +ggtitle(paste0("Patient: OV-", OV_num," ultrametric tree")) +
      theme(legend.position = "right", plot.title = element_text(hjust = 0.5), legend.key.size = unit(0.2, "cm"))  + scale_x_continuous(expand = expansion(mult = c(0,0.05))) + geom_fruit(
      data = mf_ring,
      geom = geom_bar,
      mapping = aes(y = label, x = fitness, fill = fitness),
      orientation = "y",
      stat = "identity",
      width = 0.4
    ) +
  scale_fill_gradient(low = "#93C477", high = "#1F78B4")



p_ultra <- p_ultra %<+% edge_groups + geom_tree(aes(color = group), na.rm = TRUE, linewidth = 0.2) + scale_color_manual(values = c("WGD -" = "#435274", "WGD +" = "#A13F55"))

p_ultra$data$group[is.na(p_ultra$data$group)] <- "WGD -"


p_output <- p_ultra



```



### Panel rootogram

One can reproduce the rootogram plots for all three subclonal patients using the following code section. In particular, one can adjust

```{r}
OV_num <- "075"
OV_class <- "Subclonal"
dir_name <- paste0("Results/patient results/",OV_class,"/OV-",OV_num,"/")

tree_ori <- readRDS(paste0(dir_name,"tree.rds"))
tree_ultra <- readRDS(paste0(dir_name,"ultra_tree.rds")) # `ultra_tree_des` and 



tree_ori <- label_internal_nodes(tree_ori)

y <- tree_ori$edge.length[tree_ori$edge_wgd == 1]
t <- tree_ultra$edge.length[tree_ori$edge_wgd == 1]

WGD_file <- paste0(dir_name,"/ultra-tree.csv")
WGD_Results <- read.csv(WGD_file, header = FALSE)
WGD_rates <- WGD_Results[2,]

tau <- WGD_rates*t
n   <- length(y)

k_max <- max(y)
k <- 0:k_max
O <- tabulate(y+1, nbins = k_max+1)
E <- sapply(k, function(kk){
  sum(dpois(kk,lambda = tau))
})




barplot(
  sqrt(O) - sqrt(E),
  names.arg = k,
  ylab = expression(sqrt(O[k]) - sqrt(E[k])),
  xlab = "Count k",
  main = "Poisson Mixture Rootogram",
  col = "steelblue"
)
abline(h = 0, lwd = 2)


sqrt_exp <- sqrt(E)
sqrt_obs <- sqrt(O)

obs_stat <- 2 * sum(
  ifelse(y == 0,
         0,
         y * log(y / tau)) -
    (y - tau)
)

B <- 1000
sim_stat <- numeric(B)
for (b in 1:B) {
  sim_branch <- rpois(length(tau), tau)
  
  sim_stat[b] <- 2 * sum(
    ifelse(sim_branch == 0,
           0,
           sim_branch * log(sim_branch / tau)) -
      (sim_branch -  tau)
  )
}

p_sim <- mean(sim_stat >= obs_stat)
p_sim

#svg("Andrew2025/Results/Figures/Subclonal_Fitness/OV-075-rooto-WGD-.svg", width = 6, height = 4)

plot(k, sqrt_exp,
     type = "n",
     ylim = range(c(sqrt_exp, sqrt_exp - sqrt_obs)),
     xlab = "Branch length",
     ylab = "Expected frequency",
     main = "Hanging rootogram for OV-075 WGD+ branches",
     font.lab = 2)

# expected curve
lines(k, sqrt_exp, lwd = 2)

# hanging bars
segments(x0 = k,
         y0 = sqrt_exp,
         x1 = k,
         y1 = sqrt_exp - sqrt_obs,
         lwd = 6,
         col = "#435274") # "WGD -" = "#435274", "WGD +" = "#A13F55"

abline(h = 0, lty = 2)


#dev.off()

x <- seq(-4,4,length.out = 1000)
y <- dnorm(x, mean = 0, sd = 1/4)
hist(sqrt(O) - sqrt(E), prob = TRUE)
lines(x,y,lwd = 2)


signs <- sign(sqrt_exp - sqrt_obs)
runs.test(as.factor(signs[1:length(signs)]))
```

### Supplementary Figures

#### (Figure s7) Plot the fitness prediction under alternative assumption.



```{r}
OV_num <- '075'

dir_name <- paste0("Results/patient results/Subclonal/OV-", OV_num,"/")

#tree_ori <- label_internal_nodes(tree_ori)
tree_ori$edge_wgd <- ifelse(tree_ori$edge_wgd == 1, "WGD +", "WGD -")

edge_groups <- data.frame(
  node = tree_ori$edge[,2],
  group = tree_ori$edge_wgd
)




Results_ultra <- readRDS(paste0(dir_name, "pFitness_ultra_des.rds"))
tree_ultra <- Results_ultra$phylo
mf_ultra <- Results_ultra$meanFitness[1:length(tree_ultra$tip.label)]
mp <- Results_ultra$marginalProb[1:length(tree_ultra$tip.label),]
d <- p_ori$data
tip_order <- rev(order(d$y[1:length(tree_ultra$tip.label)]))
b_rates <- 1.1*1.1^(0:15)
colnames(mp) <- c(as.character(round(b_rates - 1,2)))
u_tree <- readRDS(paste0(dir_name,"ultra_tree.rds"))
hc_tree <- as.hclust(tree_ultra)
mp_ht <- Heatmap(
  mp[tip_order, ],
  cluster_rows = hc_tree,
  cluster_columns = FALSE,
  col = c("#deebf7","#3182bd"),
  column_names_rot = 90,
  border = TRUE
)

#svg(paste0("Andrew2025/Results/Figures/Subclonal_Fitness/OV-",OV_num,"-alt-mp.svg"), width = 4, height = 6)

mp_ht
#dev.off()

mf_ring <- tibble(
  label = tree_ultra$tip.label,
  fitness = mf_ultra
)


p_ultra <- ggtree(tree_ultra, size = 0) +ggtitle(paste0("Patient: OV-", OV_num," ultrametric tree")) +
      theme(legend.position = "right", plot.title = element_text(hjust = 0.5), legend.key.size = unit(0.2, "cm"))  + scale_x_continuous(expand = expansion(mult = c(0,0.05))) + geom_fruit(
      data = mf_ring,
      geom = geom_bar,
      mapping = aes(y = label, x = fitness, fill = fitness),
      orientation = "y",
      stat = "identity",
      width = 0.4
    ) +
  scale_fill_gradient(low = "#93C477", high = "#1F78B4")



p_ultra <- p_ultra %<+% edge_groups + geom_tree(aes(color = group), na.rm = TRUE, linewidth = 0.2) + scale_color_manual(values = c("WGD -" = "#435274", "WGD +" = "#A13F55"))

p_ultra$data$group[is.na(p_ultra$data$group)] <- "WGD -"


p_output <- p_ori + p_ultra



```

#### (Figure s7) Plot subclonal rootogram (alternative WGD assumption)

```{r}

OV_num <- "075"
OV_class <- "Subclonal"
dir_name <- paste0("Results/patient results/",OV_class,"/OV-",OV_num,"/")

tree_ori <- readRDS(paste0(dir_name,"tree.rds"))
tree_ultra <- readRDS(paste0(dir_name,"ultra_tree_des.rds"))

#phy <- readRDS("Andrew2025/Results/CN_ultrametric_tree/OV-075/OV-075-WGD-tree_dec.rds") # Tree that assume WGD on descendent cell
#tree_ultra <- phy

#tree_ori <- label_internal_nodes_dec(tree_ori)

y <- tree_ori$edge.length[tree_ultra$edge_wgd == 1]
t <- tree_ultra$edge.length[tree_ultra$edge_wgd == 1]

WGD_file <- paste0("Results/patient results/",OV_class,"/OV-",OV_num,"/ultra-tree_des.csv")
WGD_Results <- read.csv(WGD_file, header = FALSE)
WGD_rates <- WGD_Results[2,]

tau <- WGD_rates*t
n   <- length(y)

k_max <- max(y)
k <- 0:k_max
O <- tabulate(y+1, nbins = k_max+1)
E <- sapply(k, function(kk){
  sum(dpois(kk,lambda = tau))
})



sqrt_exp <- sqrt(E)
sqrt_obs <- sqrt(O)

#svg("Andrew2025/Results/Figures/Subclonal_Fitness/OV-075-rooto-alt-WGD+.svg", width = 6, height = 4)

plot(k, sqrt_exp,
     type = "n",
     ylim = range(c(sqrt_exp, sqrt_exp - sqrt_obs)),
     xlab = "Branch length",
     ylab = "Expected frequency",
     main = "Hanging rootogram for OV-075 WGD branches",
     font.lab = 2)

# expected curve
lines(k, sqrt_exp, lwd = 2)

# hanging bars
segments(x0 = k,
         y0 = sqrt_exp,
         x1 = k,
         y1 = sqrt_exp - sqrt_obs,
         lwd = 6,
         col = "#A13F55") # WGD: #F8766D -> RGBA = [161 63 85 100] #A13F55, non-WGD: #00BFC4 RGBA = [67 82 116 100] #435274. Light purple (WGD alt): RGBA [162 121 143 100].

abline(h = 0, lty = 2)


#dev.off()




signs <- sign(sqrt_exp - sqrt_obs)
runs.test(as.factor(signs[1:length(signs)]))
```

#### (Figure s7) Plot Subclonal rootogram (Negative Binomial)

```{r}
OV_num <- "075"
OV_class <- "Subclonal"
dir_name <- paste0("Results/patient results/",OV_class,"/OV-",OV_num,"/")

tree_ori <- readRDS(paste0(dir_name,"tree.rds"))
tree_ultra <- readRDS(paste0(dir_name,"ultra_tree.rds"))

#phy <- readRDS("Andrew2025/Results/CN_ultrametric_tree/OV-075/OV-075-WGD-tree_dec.rds") # Tree that assume WGD on descendent cell
#tree_ultra <- phy

tree_ori <- label_internal_nodes(tree_ori)

y <- tree_ori$edge.length[tree_ori$edge_wgd == 1]
t <- tree_ultra$edge.length[tree_ori$edge_wgd == 1]

WGD_file <- paste0("Results/patient results/",OV_class,"/OV-",OV_num,"/ultra-tree_nb.csv")
WGD_Results <- read.csv(WGD_file, header = FALSE)
WGD_rates <- WGD_Results[2,]
r         <- WGD_Results[dim(WGD_Results)[1],]

p <- r/(r + WGD_rates*t)

k_max <- max(y)
k <- 0:k_max
O <- tabulate(y+1, nbins = k_max+1)
E <- sapply(k, function(kk){
  sum(dnbinom(kk,size = r, prob = p))
})



sqrt_exp <- sqrt(E)
sqrt_obs <- sqrt(O)

#svg("Andrew2025/Results/Figures/Subclonal_Fitness/OV-075-rooto-nb-WGD+.svg", width = 6, height = 4)

plot(k, sqrt_exp,
     type = "n",
     ylim = range(c(sqrt_exp, sqrt_exp - sqrt_obs)),
     xlab = "Branch length",
     ylab = "Expected frequency",
     main = "Hanging rootogram for OV-075 WGD branches",
     font.lab = 2)

# expected curve
lines(k, sqrt_exp, lwd = 2)

# hanging bars
segments(x0 = k,
         y0 = sqrt_exp,
         x1 = k,
         y1 = sqrt_exp - sqrt_obs,
         lwd = 6,
         col = "#A13F55") # WGD: #A13F55, non-WGD: #435274.

abline(h = 0, lty = 2)


#dev.off()




signs <- sign(sqrt_exp - sqrt_obs)
runs.test(as.factor(signs))
```






