Script description:

This script aims to streamline the spatial in silico analysis for LeafRank.
The backbone for this script is plotting the main figures in manuscript, i.e. part of Figure 4(In silico WGD Robustness).




### Initialization

```{r}
library('reticulate')
library("ape")
library("tibble")
library('readr')
library("ggplot2")
library("ggtreeExtra")
library("ggpubr")
library('here')
library('ggtree')
library('phangorn')
library('tidyr')
library('dplyr')
library('gridExtra')
library('patchwork')

source('../R/fitness_inference.R')
source('../R/tree_utils.R')


```




## Figure 4(In silico WGD robustness)




### Collecting important metrics 

The following scripts are used to collect all the correlation among 50 (simulation) times 50 (resampling) results.
Besides, several metrics are computed:
1. maximum fitness
2. variance of fitness


```{r}
num_batch <- 50

corr_true_TT <- matrix(0, 50, 50, byrow = FALSE)
corr_true_WGD <- matrix(0, 50, 50, byrow = FALSE)
corr_true_DM  <- matrix(0, 50, 50, byrow = FALSE)
corr_TT_WGD  <- matrix(0, 50, 50, byrow = FALSE)
max_fitness <- matrix(0, 50, 50, byrow = FALSE)
max_fitness_TT <- matrix(0, 50, 50, byrow = FALSE)
max_fitness_DM <- matrix(0, 50, 50, byrow = FALSE)
max_fitness_WGD <- matrix(0, 50, 50, byrow = FALSE)
var_fitness <- matrix(0, 50, 50, byrow = FALSE)
wrong_WGD <- matrix(0, 50, 50, byrow = FALSE)
wrong_DM  <- matrix(0, 50, 50, byrow = FALSE)

for (s_idx in 1:50){


  sim_num <- as.character(s_idx)
  print(sim_num)




    batch_num <- as.character(i)
    dir_name <- paste0("Results/spatial simulation results/SIM-",sim_num,"/")
    
    
    Results_TT <- tryCatch({readRDS(paste0(dir_name,'P_fitness_TT.out.rds'))}, warning = function(w){
      return(NULL)
    })
    Results_WGD <- tryCatch({readRDS(paste0(dir_name,'P_fitness_WGD.out.rds'))}, warning = function(w){
      return(NULL)
    })
    Results_DM <- tryCatch({readRDS(paste0(dir_name,'P_fitness_DM.out.rds'))}, warning = function(w){
      return(NULL)
    })
    if (!is.null(Results_TT)){
      phy_TT  <- Results_TT$phylo
      mf_TT <- Results_TT$meanFitness[1:length(phy_TT$tip.label)]
      Fitness_true    <- as.numeric(phy_TT$fitness)
      corr_true_TT[s_idx,i]<- cor(mf_TT,Fitness_true,method='spearman')
      max_fitness[s_idx,i] <- max(Fitness_true)
      max_fitness_TT[s_idx,i] <- max(mf_TT)
      var_fitness[s_idx,i] <- var(Fitness_true)
    }else{next}
    if (!is.null(Results_DM)){
      phy_DM  <- Results_DM$phylo
      mf_DM <- Results_DM$meanFitness[1:length(phy_DM$tip.label)]
      corr_true_DM[s_idx,i] <- cor(mf_DM,Fitness_true,method='spearman')
      max_fitness_DM[s_idx,i] <- max(mf_DM)
      if (length(phy_DM$fitness)>0) wrong_DM[s_idx,i] <-1
    }
    
    if (!is.null(Results_WGD)){
      phy_WGD  <- Results_WGD$phylo
      mf_WGD <- Results_WGD$meanFitness[1:length(phy_WGD$tip.label)]
      corr_true_WGD[s_idx,i] <- cor(mf_WGD,Fitness_true,method='spearman')
      max_fitness_WGD[s_idx,i] <- max(mf_WGD)
      corr_TT_WGD[s_idx,i] <- cor(mf_WGD,mf_TT, method = 'spearman')
      if (length(phy_WGD$fitness)>0) wrong_WGD[s_idx,i] <-1
    }
    
    
    
    
    
  
  





}



```

```{r}
corr_true_TT_tot <- corr_true_TT[corr_true_TT != 0]
corr_true_DM_tot <- corr_true_DM[corr_true_DM != 0]
corr_true_WGD_tot <- corr_true_WGD[corr_true_WGD != 0]
sample_mean_TT <- apply(corr_true_TT,1,function(x)mean(x[x!=0]))
sample_var_TT  <- apply(corr_true_TT,1,function(x)var(x[x!=0]))
sample_mean_DM <- apply(corr_true_DM,1,function(x)mean(x[x!=0]))
sample_var_DM  <- apply(corr_true_DM,1,function(x)var(x[x!=0]))
sample_mean_WGD <- apply(corr_true_WGD,1,function(x)mean(x[x!=0]))
sample_var_WGD  <- apply(corr_true_WGD,1,function(x)var(x[x!=0]))
max_fitness_tot <- max_fitness[max_fitness != 0]
var_fitness_tot <- var_fitness[var_fitness != 0]

p <- boxplot(list(TT = corr_true_TT_tot, DM = corr_true_DM_tot, WGD = corr_true_WGD_tot))

p_mean_hist <- hist(sample_mean_TT)

```


### Panel C

The illustrated example is plotted as "p" here.

```{r}

s_idx <- 8
sim_num <- as.character(s_idx)
print(sim_num)
#for (i in 1:50){
  i <- 1
  case_num <- "8"
  batch_num <- as.character(i)
  
  dir_name <- paste0("Results/spatial simulation results/SIM-",sim_num,"/")
    
    
    Results_TT <- tryCatch({readRDS(paste0(dir_name,'P_fitness_TT.out.rds'))}, warning = function(w){
      return(NULL)
    })
    Results_WGD <- tryCatch({readRDS(paste0(dir_name,'P_fitness_WGD.out.rds'))}, warning = function(w){
      return(NULL)
    })
    Results_DM <- tryCatch({readRDS(paste0(dir_name,'P_fitness_DM.out.rds'))}, warning = function(w){
      return(NULL)
    })
  phy_TT <- Results_TT$phylo
  phy_DM <- Results_DM$phylo
  phy_WGD <- Results_WGD$phylo
  phy_TT$tip.label <- phy_DM$tip.label
  phy_WGD_o <- readRDS(paste0(dir_name,"Tree-WGD.rds"))
  tip_wgd <- phy_WGD_o$node_wgd[1:250]

  Fitness_true    <- as.numeric(phy_TT$fitness)
  mf_TT <- Results_TT$meanFitness[1:length(phy_TT$tip.label)]
  mf_DM <- Results_DM$meanFitness[1:length(phy_DM$tip.label)]
  mf_WGD <- Results_WGD$meanFitness[1:length(phy_WGD$tip.label)]

  wgd_group <- ifelse(tip_wgd == 0, "WGD -", "WGD +")
  
  groups <- data.frame(
    label = phy_TT$tip.label,
    group = wgd_group
  )
  otu_groups <- split(groups$label, groups$group)
  tree_grouped <- groupOTU(phy_TT, otu_groups)
  tree_DM_grouped <- groupOTU(phy_DM, otu_groups)
  tree_WGD_grouped <- groupOTU(phy_WGD, otu_groups)
  
  phy_TT$node_wgd <- c(tip_wgd, rep(-1, phy_TT$Nnode))
  phy_DM$node_wgd <- c(tip_wgd, rep(-1, phy_DM$Nnode))
  phy_WGD$node_wgd <- c(tip_wgd, rep(-1, phy_WGD$Nnode))
  phy_TT <- label_internal_nodes(phy_TT)
  phy_DM <- label_internal_nodes(phy_DM)
  phy_WGD <- label_internal_nodes(phy_WGD)
  
  phy_TT$edge_wgd <- ifelse(phy_TT$edge_wgd == 0, "WGD -", "WGD +")
  
  edge_group_TT <- data.frame(
    node = phy_TT$edge[,2],
    group = phy_TT$edge_wgd
  )
  
  
  phy_DM$edge_wgd <- ifelse(phy_DM$edge_wgd == 0, "WGD -", "WGD +")
  
  edge_group_DM <- data.frame(
    node = phy_DM$edge[,2],
    group = phy_DM$edge_wgd
  )
  
  phy_WGD$edge_wgd <- ifelse(phy_WGD$edge_wgd == 0, "WGD -", "WGD +")
  
  edge_group_WGD <- data.frame(
    node = phy_WGD$edge[,2],
    group = phy_WGD$edge_wgd
  )
  
  
  Tf_ring <- tibble(
    label = phy_TT$tip.label,
    fitness = Fitness_true
  )
  
  
  TT_ring <- tibble(
    label = phy_TT$tip.label,
    fitness = mf_TT
  )
  
  DM_ring <- tibble(
    label = phy_DM$tip.label,
    fitness = mf_DM
  )
  
  WGD_ring <- tibble(
    label = phy_WGD$tip.label,
    fitness = mf_WGD
  )
  
  
  p_Tf <- ggtree(phy_TT, layout = 'circular', size = 0)+ 
    ggtitle("True ranking") +
    theme(plot.title = element_text(size = 14, vjust = -2.5,hjust = 0.5, face = "bold"), 
          legend.position = "bottom",
          legend.key.size = unit(0.3, "cm"),
          legend.title = element_text(size = 14, face = "bold"),
          legend.text = element_text(size = 12, face = "bold")
          ) + 
    geom_fruit(
        data = Tf_ring,
        geom = geom_bar,
        mapping = aes(y = label, x = fitness, fill = fitness),
        orientation = "y",
        stat = "identity",
        width = 0.3
      ) +
    scale_fill_gradient(low = "#93C477", high = "#1F78B4") + 
    guides(fill = guide_colorbar(order = 1,
                                 barwidth = unit(6, "cm"),
                                 barheight = unit(0.4,"cm"),
                                 title.position = "top"), 
           color = guide_legend(order = 2,
                                title.position = "top"))
  
  p_Tf <- p_Tf %<+% edge_group_TT + geom_tree(aes(color = group), na.rm = TRUE,linewidth = 0.3) + scale_color_manual(values = c("WGD -" = "#435274", "WGD +" = "#A13F55"))
  
  

  
  
  p_TT <- ggtree(phy_TT, layout = 'circular', size = 0)+
    ggtitle(paste0("TT ranking (rho: ",format(round(cor(mf_TT,Fitness_true,method = 'spearman'), 4), nsmall = 4),")")) +
    theme(plot.title = element_text(size = 14, vjust = -2.5,hjust = 0.5, face = "bold"), 
          legend.position = "bottom",
          legend.key.size = unit(0.3, "cm"),
          legend.title = element_text(size = 14, face = "bold"),
          legend.text = element_text(size = 12, face = "bold")
          ) + 
    geom_fruit(
        data = TT_ring,
        geom = geom_bar,
        mapping = aes(y = label, x = fitness, fill = fitness),
        orientation = "y",
        stat = "identity",
        width = 0.3
      ) +
    scale_fill_gradient(low = "#93C477", high = "#1F78B4") + 
    guides(fill = guide_colorbar(order = 1,
                                 barwidth = unit(6, "cm"),
                                 barheight = unit(0.4,"cm"),
                                 title.position = "top"), 
           color = guide_legend(order = 2, 
                                title.position = "top"))
  p_TT <- p_TT %<+% edge_group_TT + geom_tree(aes(color = group), na.rm = TRUE,linewidth = 0.3) + scale_color_manual(values = c("WGD -" = "#435274", "WGD +" = "#A13F55"))
  
  rng <- range(DM_ring$fitness, na.rm = TRUE)
  b_rng <- c(rng[1] + 0.05*diff(rng), rng[2] - 0.05*diff(rng))
  
  p_DM <- ggtree(phy_DM, layout = 'circular', size = 0)+
    ggtitle(paste0("DM ranking (rho: ",format(round(cor(mf_DM,Fitness_true,method = 'spearman'), 4), nsmall = 4),")")) +
    theme(plot.title = element_text(size = 14, vjust = -2.5,hjust = 0.5, face = "bold"), 
          legend.position = "bottom",
          legend.key.size = unit(0.3, "cm"),
          legend.title = element_text(size = 14, face = "bold"),
          legend.text = element_text(size = 12, face = "bold")
          ) + 
    geom_fruit(
        data = DM_ring,
        geom = geom_bar,
        mapping = aes(y = label, x = fitness, fill = fitness),
        orientation = "y",
        stat = "identity",
        width = 0.3
      ) +
    scale_fill_gradient(low = "#93C477", high = "#1F78B4",
                        limits = rng,
                        breaks = b_rng,
                        labels = c("Low","High"),
                        name = "Fitness") + 
    guides(fill = guide_colorbar(order = 1,
                                 barwidth = unit(6, "cm"),
                                 barheight = unit(0.4,"cm"),
                                 title.position = "top"), 
           color = guide_legend(order = 2,
                                title.position = "top"))
  p_DM <- p_DM %<+% edge_group_DM + geom_tree(aes(color = group), na.rm = TRUE, linewidth = 0.3) + scale_color_manual(values = c("WGD -" = "#435274", "WGD +" = "#A13F55"))
  
  p_WGD <- ggtree(phy_WGD, layout = 'circular', size = 0)+
    ggtitle(paste0("WGD ranking (rho: ",format(round(cor(mf_WGD,Fitness_true,method = 'spearman'), 4), nsmall = 4),")")) +
    theme(plot.title = element_text(size = 14, vjust = -2.5,hjust = 0.5, face = "bold"), 
          legend.position = "bottom",
          legend.key.size = unit(0.3, "cm"),
          legend.title = element_text(size = 14, face = "bold"),
          legend.text = element_text(size = 12, face = "bold")
          ) + 
    geom_fruit(
        data = WGD_ring,
        geom = geom_bar,
        mapping = aes(y = label, x = fitness, fill = fitness),
        orientation = "y",
        stat = "identity",
        width = 0.3
      ) +
    scale_fill_gradient(low = "#93C477", high = "#1F78B4") + 
    guides(fill = guide_colorbar(order = 1,
                                 barwidth = unit(6, "cm"),
                                 barheight = unit(0.4,"cm"),
                                 title.position = "top"), 
           color = guide_legend(order = 2,
                                title.position = "top"))
  p_WGD <- p_WGD %<+% edge_group_WGD + geom_tree(aes(color = group), na.rm = TRUE, linewidth = 0.3, ladderize = FALSE) + scale_color_manual(values = c("WGD -" = "#435274", "WGD +" = "#A13F55"))
  
  
  
  
  shard_legend <- cowplot::get_legend(p_DM)
  
  p_Tf_n <- p_Tf + theme(legend.position = "none")
  p_TT_n <- p_TT + theme(legend.position = "none")
  p_DM_n <- p_DM + theme(legend.position = "none")
  p_WGD_n <- p_WGD + theme(legend.position = "none")
  
  
#  p <- (p_Tf|p_mf) + 
#    plot_annotation(title = paste0("Spearman correlation: ", corr_true_ori[i])) &
#    theme(plot.title = element_text(hjust = 0.5))

#}
  title_grob <- textGrob(
    "Non-Spatial Simulation Analysis",
    gp = gpar(fontface = "bold", fontsize = 16),
    just = "center")
  
  p <- grid.arrange(
    arrangeGrob(p_Tf_n,p_TT_n,p_DM_n,p_WGD_n, nrow = 1), 
    shard_legend,
    title_grob,
    heights = c(0.5,0.1,0.1),
    ncol = 1
  )

  #ggsave("/projects/standard/ruping/shared/wu000766/Fitness_ranking/Simulated_COMET/wgd/Figures/Fitness_plot.svg", plot = p, width = 16, height = 4)
#}
```


### Panel E

Both WGD- and WGD+ rootogram are plotted here.

```{r}
s_idx <- 8
sim_num <- as.character(s_idx)
print(sim_num)
#for (i in 1:50){
  i <- 1
  case_num <- "8"
  batch_num <- as.character(i)
  
  dir_name <- paste0("Results/spatial simulation results/SIM-",sim_num,"/")
    DM_name <- paste0(dir_name,"Tree-WGD.rds") # Note this is the original WGD tree before ultrametric tree conversion.
    TT_name <- paste0(dir_name,"Tree-TT.rds")
    # Here we use true total time tree to determine the true overall time horizon
    
    
    
    
    #Ori_tree <- read.tree(name)
    #Ori_tree <- readRDS(name)
    DM_tree  <- readRDS(DM_name)
    TT_tree  <- readRDS(TT_name)
    WGD_file <- paste0(dir_name,"ultra-tree.csv")
    
    #discrete_tree <- readRDS(discrete_file)
    #WGD_R_tree <- readRDS(WGD_R_file)
    WGD_Results <- read.csv(WGD_file, header = FALSE)
    WGD_Results <- WGD_Results[,1]
    WGD_tree <- DM_tree
    e1 <- WGD_tree$edge[,1]
    e2 <- WGD_tree$edge[,2]
    li <- length(WGD_Results)
    
    WGD_ages <- c(rep(0,li),1,WGD_Results[3:length(WGD_Results)])
    WGD_tree$edge.length <- WGD_ages[e1] - WGD_ages[e2]
    

y <- DM_tree$edge.length[DM_tree$edge_wgd == 0]
t <- WGD_tree$edge.length[DM_tree$edge_wgd == 0]


WGD_rates <- WGD_Results[1]

tau <- WGD_rates*t
n   <- length(y)

k_max <- max(y)
k <- 0:k_max
O <- tabulate(y+1, nbins = k_max+1)
E <- sapply(k, function(kk){
  sum(dpois(kk,lambda = tau))
})



#svg("Simulated_COMET/wgd/Figures/SIM_rooto.svg", width = 9, height = 4)

par(mfrow = c(1,2), oma = c(0,0,3,0))

y <- DM_tree$edge.length[DM_tree$edge_wgd == 0]
t <- WGD_tree$edge.length[DM_tree$edge_wgd == 0]


WGD_rates <- WGD_Results[1]

tau <- WGD_rates*t
n   <- length(y)

k_max <- max(y)
k <- 0:k_max
O <- tabulate(y+1, nbins = k_max+1)
E <- sapply(k, function(kk){
  sum(dpois(kk,lambda = tau))
})

sqrt_exp <- sqrt(E)
sqrt_obs <- sqrt(O)

signs <- sign(sqrt_exp - sqrt_obs)
signs_nWGD<- runs.test(as.factor(signs))

plot(k, sqrt_exp,
     type = "n",
     ylim = range(c(sqrt_exp, sqrt_exp - sqrt_obs)),
     xlab = "WGD- Branch Length",
     ylab = "\u221AFrequency",
     main = "",
     font.lab = 2)

# expected curve
lines(k, sqrt_exp, lwd = 2)

# hanging bars
segments(x0 = k,
         y0 = sqrt_exp,
         x1 = k,
         y1 = sqrt_exp - sqrt_obs,
         lwd = 3,
         col = "#435274") # WGD: #F8766D, non-WGD: #00BFC4

abline(h = 0, lty = 2)

legend("topright",
       legend = c("Expected","Observed"),
       col = c("black","#435274"),
       lty = 1,
       lwd = 3,
       bty = "n")


y <- DM_tree$edge.length[DM_tree$edge_wgd == 1]
t <- WGD_tree$edge.length[DM_tree$edge_wgd == 1]


WGD_rates <- WGD_Results[2]

tau <- WGD_rates*t
n   <- length(y)

k_max <- max(y)
k <- 0:k_max
O <- tabulate(y+1, nbins = k_max+1)
E <- sapply(k, function(kk){
  sum(dpois(kk,lambda = tau))
})

sqrt_exp <- sqrt(E)
sqrt_obs <- sqrt(O)

plot(k, sqrt_exp,
     type = "n",
     ylim = range(c(sqrt_exp, sqrt_exp - sqrt_obs)),
     xlab = "WGD+ Branch Length",
     ylab = "",
     font.lab = 2)

# expected curve
lines(k, sqrt_exp, lwd = 2)

# hanging bars
segments(x0 = k,
         y0 = sqrt_exp,
         x1 = k,
         y1 = sqrt_exp - sqrt_obs,
         lwd = 3,
         col = "#A13F55") # WGD: #F8766D, non-WGD: #00BFC4

abline(h = 0, lty = 2)


mtext("Hanging Rootogram for Spatial Simulation",
      side = 3, outer = TRUE, line = 1, cex = 1.3, font = 2)

legend("topright",
       legend = c("Expected","Observed"),
       col = c("black","#A13F55"),
       lty = 1,
       lwd = 3,
       bty = "n")




#dev.off()




signs <- sign(sqrt_exp - sqrt_obs)
signs_WGD <- runs.test(as.factor(signs))
```


### Panel H

We then examine the relationship between sampling variability and inference accuracy. 
Two types of sampling are consider:
1. Sample from original population
2. Subsample from sampled tree

#### Collecting all sampling inference results

```{r}
#corr_true_re <- matrix(0, nrow = 50, ncol = 50)
corr_true_ori <- rep(0,50)
corr_ori_sub <- matrix(0, nrow = 50, ncol = 50)
corr_ori_re_sub <- matrix()


for (ii in 1:50){
  batch_num <- as.character(ii)
  dir_name <- paste0("Results/spatial simulation results/SIM-",batch_num,"/")
  TT_name <- paste0(dir_name,"P_fitness_TT.out.rds")
  Results_TT <- readRDS(TT_name)
  #phy_TT <- readRDS(TT_name)
  phy_TT <- Results_TT$phylo
  mf_TT <- Results_TT$meanFitness[1:length(phy_TT$tip.label)]
  Fitness_true <- as.numeric(phy_TT$fitness)
  corr_true_ori[ii] <- cor(mf_TT,Fitness_true,method='spearman')
  for (jj in 1:50){
    sample_run_num <- as.character(jj)
    
    
    
    dir_name_sub <- paste0("Results/spatial simulation results/SIM-",batch_num,"/Subsampling/")
    TT_name_sub <- paste0(dir_name_sub,"P_fitness-",sample_run_num,".out.rds")
    Results_TT_sub <- readRDS(TT_name_sub)
    phy_TT_sub <- Results_TT_sub$phylo
    index_sub <- match(as.numeric(phy_TT_sub$tip.label), phy_TT$tip.label)
    mf_TT_sub <- Results_TT_sub$meanFitness[1:length(phy_TT_sub$tip.label)]
    
    
    corr_ori_sub[jj,ii] <- cor(mf_TT_sub,mf_TT[index_sub],method='spearman')
    
    
    
    
  }
}

```


#### Visualization 

```{r}






sub_df <- data.frame(
  idx = 1:50,
  Acc = corr_true_ori,
  sub_var = apply(corr_ori_sub,2,var)
)

sub_df$color <- "black"

cor_test <- cor.test(sub_df$sub_var, sub_df$Acc)
label_text <- paste0(
  "Pearson = ", round(cor_test$estimate, 2),
  "\nP = ", signif(cor_test$p.value, 2)
)

p_sub2 <- ggplot(sub_df, aes(x = Acc, y = sub_var)) +
  geom_point(aes(color = color), size = 2.5)+
  scale_color_identity()+
  geom_smooth(method = "lm", se = TRUE) +
  annotate(
    "text",
    x = max(sub_df$Acc) * 0.8,
    y = max(sub_df$sub_var) * 0.9,
    label = label_text,
    size = 5
  ) +
  labs(
    y = "Subsampling Variance",
    x = "Prediction Accuracy (Spearman)"
  ) +
  theme_bw()+
  theme(
    axis.text = element_text(size = 14),
    plot.title = element_text(size = 18, face = "bold")
  ) + scale_y_log10()

#ggsave("MATLAB/Simulation/Figures/Var_Acc_Re.svg",plot = p_re, width = 7, height = 4)
#ggsave("MATLAB/Simulation/Figures/Var_Acc_Sub.svg",plot = p_sub, width = 7, height = 4)


```



