Script description:

This script aims to streamline the nonspatial in silico analysis for LeafRank.
The backbone for this script is plotting the main figures in manuscript, i.e. Figure 2(Base in silico experiment), Figure 3(In silico Parameter robustness), and part of Figure 4(In silico WGD Robustness).
The majority data for non-spatial simulation are stored at "MATLAB/Simulation/Results".
One can find the simulation setting in "MATLAB/Simulation/Simulation_description.txt".

## Initialization
```{r}
library('reticulate')
library("ape")
library("tibble")
library(tseries)
library(ComplexHeatmap)
library('readr')
library("ggplot2")
library("ggtreeExtra")
library("ggpubr")
library('ggtree')
library('phangorn')
library('tidyr')
library('dplyr')
library('gridExtra')
library('patchwork')
library('R.matlab')


source('../R/fitness_inference.R')
source('../R/tree_utils.R')
#source(here('chronos_WGD.R'))

```


## Figure 2(Base in silico experiment)

### Robustness w.r.t. simulation (Panel A data)

We collected the inference accuracy as correlation from all 50 synthesis tumor.

```{r}
num_sim <- 50

corr_true_ori <- rep(0,num_sim)
max_Tf <- rep(0,num_sim)
is_Lowest <- rep(0,num_sim)
acc <- rep(0,num_sim)
b_rates <- 0.2*1.2^(0:7)

for (i in 1:num_sim){
  batch_num <- as.character(i)
  dir_name <- paste0("Results/non-spatial simulation results/SIM_",batch_num)
  TT_name <- paste0(dir_name,"/P_fitness.out.rds")
  Results_TT <- readRDS(TT_name)
  
  phy_ori <- readRDS(paste0(dir_name,"/Tree.rds"))
  Fitness_true    <- as.numeric(phy_ori$fitness$V1)
  max_Tf[i] <- max(Fitness_true)
  
  #Low_idx <- Fitness_true == min(Fitness_true)
  
  
  mf_TT <- Results_TT$meanFitness[1:length(phy_ori$tip.label)]
  
  iL <- wilcox.test(mf_TT[Fitness_true == min(Fitness_true)], mf_TT[Fitness_true == min(Fitness_true) + 1], alternative = "less")
  is_Lowest[i] <- iL$p.value
  
  corr_true_ori[i]<- cor(mf_TT,Fitness_true,method='spearman')
  acc[i] <-sum(abs(mf_TT-b_rates[Fitness_true])/b_rates[Fitness_true])/length(Fitness_true)
}

results <- data.frame(
  replicate = 1:50,
  corr = corr_true_ori
)
p_result <- ggplot(results, aes(x = corr)) +
  geom_histogram(
    bins = 20,
    fill = "#1F78B4",
    color = "white",
    alpha = 0.8
  ) +
  labs(
    title = "Robustness of Ranking",
    x = "Spearman correlation with true ranking",
    y = "Frequency"
  ) +
  theme_minimal(base_size = 14)



```

### Plotting overall correlation (Panel A plot)

```{r}

plot_df <- data.frame(
  idx    = 1:50,
  Spearman = corr_true_ori
)
plot_df$color <- ifelse(plot_df$idx == 24, "#1F78B4",
                        ifelse(plot_df$idx == 35, "#A13F55","black"))



p <- ggplot(plot_df, aes(x = "All synthetic tumors", y = Spearman)) +
  geom_violin(fill = "lightgrey") +
  geom_jitter(aes(color = color), width = 0.05,  size = 3)+
  scale_color_identity()+
  theme_classic()


```


### Plotting Illustrative examples (Panel B plot)

We compared the illustrative example 24(good) versus 35(bad). The batch index can be modified to check other examples.

```{r}


batch_num <- as.character(24)
  
  dir_name <- paste0("Results/non-spatial simulation results/SIM_",batch_num)
  TT_name <- paste0(dir_name,"/P_fitness.out.rds")
  Results_TT <- readRDS(TT_name)
  
  phy_TT <- readRDS(paste0(dir_name,"/Tree.rds"))
  Fitness_true    <- as.numeric(phy_TT$fitness$V1)
  mf_TT <- Results_TT$meanFitness[1:length(phy_TT$tip.label)]
  
  
  Tf_ring <- tibble(
    label = phy_TT$tip.label,
    fitness = Fitness_true
  )
  
  
  TT_ring <- tibble(
    label = phy_TT$tip.label,
    fitness = mf_TT
  )
  
  
  
  
  p_Tf_24 <- ggtree(phy_TT, layout = 'circular', size = 0.3)+ labs(title = "True ranking") +
    theme(legend.position = "right", plot.title = element_text(hjust = 0.5), legend.key.size = unit(0.2, "cm"), plot.margin = unit(c(0,0,0,0), "cm")) + 
    geom_fruit(
        data = Tf_ring,
        geom = geom_bar,
        mapping = aes(y = label, x = fitness, fill = fitness),
        orientation = "y",
        stat = "identity",
        width = 0.3
      ) +
    scale_fill_gradient(low = "#93C477", high = "#1F78B4") + guides(fill = guide_colorbar(order = 1), color = guide_legend(order = 2)) # skyblue - darkblue.
  
  

  
  
  p_TT_24 <- ggtree(phy_TT, layout = 'circular', size = 0.3)+labs(title = paste0("TT ranking (rho: ",format(round(corr_true_ori[24], 4), nsmall = 4),")")) +
    theme(legend.position = "right", plot.title = element_text(hjust = 0.5), legend.key.size = unit(0.2, "cm"), plot.margin = unit(c(0,0,0,0), "cm")) + 
    geom_fruit(
        data = TT_ring,
        geom = geom_bar,
        mapping = aes(y = label, x = fitness, fill = fitness),
        orientation = "y",
        stat = "identity",
        width = 0.3
      ) +
    scale_fill_gradient(low = "#93C477", high = "#1F78B4")+ guides(fill = guide_colorbar(order = 1), color = guide_legend(order = 2))


  
batch_num <- as.character(35)
  
  dir_name <- paste0("Results/non-spatial simulation results/SIM_",batch_num)
  TT_name <- paste0(dir_name,"/P_fitness.out.rds")
  Results_TT <- readRDS(TT_name)
  
  phy_TT <- readRDS(paste0(dir_name,"/Tree.rds"))
  Fitness_true    <- as.numeric(phy_TT$fitness$V1)
  mf_TT <- Results_TT$meanFitness[1:length(phy_TT$tip.label)]
  
  
  Tf_ring <- tibble(
    label = phy_TT$tip.label,
    fitness = Fitness_true
  )
  
  
  TT_ring <- tibble(
    label = phy_TT$tip.label,
    fitness = mf_TT
  )
  
  
  
  
  p_Tf_35 <- ggtree(phy_TT, layout = 'circular', size = 0.3)+ labs(title = "True ranking") +
    theme(legend.position = "right", plot.title = element_text(hjust = 0.5), legend.key.size = unit(0.2, "cm"), plot.margin = unit(c(0,0,0,0), "cm")) + 
    geom_fruit(
        data = Tf_ring,
        geom = geom_bar,
        mapping = aes(y = label, x = fitness, fill = fitness),
        orientation = "y",
        stat = "identity",
        width = 0.3
      ) +
    scale_fill_gradient(low = "#93C477", high = "#1F78B4") + guides(fill = guide_colorbar(order = 1), color = guide_legend(order = 2)) 
  
  

  
  
  p_TT_35 <- ggtree(phy_TT, layout = 'circular', size = 0.3)+labs(title = paste0("TT ranking (rho: ",format(round(corr_true_ori[35], 4), nsmall = 4),")")) +
    theme(legend.position = "right", plot.title = element_text(hjust = 0.5), legend.key.size = unit(0.2, "cm"), plot.margin = unit(c(0,0,0,0), "cm")) + 
    geom_fruit(
        data = TT_ring,
        geom = geom_bar,
        mapping = aes(y = label, x = fitness, fill = fitness),
        orientation = "y",
        stat = "identity",
        width = 0.3
      ) +
    scale_fill_gradient(low = "#93C477", high = "#1F78B4")+ guides(fill = guide_colorbar(order = 1), color = guide_legend(order = 2))


p <- grid.arrange(p_Tf_24, p_TT_24, p_Tf_35, p_TT_35, nrow = 2, ncol = 2)

```



We examine the stability of LeafRank inference in two aspect, where the potential randomness introduced.
1. Randomness introduced during the simulation.
2. Randomness introduced during the sampling.



### Robustness w.r.t. sampling (Panel C data)

We collected the inference accuracy from all 100 re-sampled TT trees for tumor #20.

```{r}
corr_sub_ori_24 <- rep(0,50)
corr_sub_ori_35 <- rep(0,50)

for (ii in 1:50){
  case_num <- "8"
  batch_num <- "24"
  sample_num <- as.character(ii)
  TT_name_ori <- paste0("Results/non-spatial simulation results/SIM_", batch_num,"/P_fitness.out.rds")
  Results_TT <- readRDS(TT_name_ori)
  phy_TT <- Results_TT$phylo
  mf_TT <- Results_TT$meanFitness[1:length(phy_TT$tip.label)]
    
  dir_name_sub <- paste0("Results/non-spatial simulation results/SIM_", batch_num,"/Subsampling/")
  TT_name_sub <- paste0(dir_name_sub,"P_fitness-",case_num,"-",batch_num,"_",sample_num,".out.rds")
  Results_TT_sub <- readRDS(TT_name_sub)
  phy_TT_sub <- Results_TT_sub$phylo
  index_sub <- as.numeric(phy_TT_sub$tip.label)
  mf_TT_sub <- Results_TT_sub$meanFitness[1:length(phy_TT_sub$tip.label)]
  corr_sub_ori_24[ii] <- cor(mf_TT_sub,mf_TT[index_sub],method='spearman')
  
  case_num <- "8"
  batch_num <- "35"
  sample_num <- as.character(ii)
  TT_name_ori <- paste0("Results/non-spatial simulation results/SIM_", batch_num,"/P_fitness.out.rds")
  Results_TT <- readRDS(TT_name_ori)
  phy_TT <- Results_TT$phylo
  mf_TT <- Results_TT$meanFitness[1:length(phy_TT$tip.label)]
    
  dir_name_sub <- paste0("Results/non-spatial simulation results/SIM_", batch_num,"/Subsampling/")
  TT_name_sub <- paste0(dir_name_sub,"P_fitness-",case_num,"-",batch_num,"_",sample_num,".out.rds")
  Results_TT_sub <- readRDS(TT_name_sub)
  phy_TT_sub <- Results_TT_sub$phylo
  index_sub <- as.numeric(phy_TT_sub$tip.label)
  mf_TT_sub <- Results_TT_sub$meanFitness[1:length(phy_TT_sub$tip.label)]
  corr_sub_ori_35[ii] <- cor(mf_TT_sub,mf_TT[index_sub],method='spearman')
}



```

### Plotting subsampling results (C plots)

```{r}
df <- data.frame(
  value = c(corr_sub_ori_24,corr_sub_ori_35),
  group = rep(c("High accuracy","Low accuracy"), each = 50)
)
p <- ggplot(df, aes(x = value, fill = group)) +
  geom_histogram(
    binwidth = 0.018,
    position = "identity"
  )+ scale_x_continuous(
    name = "Spearman correlation",
    limits = c(-0.05,1.01),
    breaks = c( 0, 0.2, 0.4, 0.6, 0.8, 1),
labels = c(
  "0\n(None)", 
  "0.2\n(weak+)", 
  "0.4\n(moderate+)", 
  "0.6\n(strong+)", 
  "0.8\n(very strong+)", 
  "1\n(perfect+)"
))+ 
  labs(title = "Robustness under sub-sampling", y = "Frequency")+
  scale_fill_manual(values = c("High accuracy" = "#425274","Low accuracy" = "#A13F55"))+
  theme_minimal()



```




We then examine the relationship between sampling variability and inference accuracy. 
Two types of sampling are consider:
1. Sample from original population
2. Subsample from sampled tree

### Collecting all sampling inference results

```{r}
corr_true_re <- matrix(0, nrow = 50, ncol = 50)
corr_true_ori <- rep(0,50)
corr_ori_sub <- matrix(0, nrow = 50, ncol = 50)
corr_ori_re_sub <- matrix()

case_num <- "8"

for (ii in 1:50){
  batch_num <- as.character(ii)
  for (jj in 1:50){
    sample_num <- as.character(jj)
    
    # dir_name_re <- paste0("MATLAB/Simulation/Results/SIM_",case_num,"_",batch_num,"/Resampling/")
    # TT_name_re <- paste0(dir_name_re,"P_fitness-",case_num,"-",batch_num,"_",sample_num,".out.rds")
    # Results_TT_re <- readRDS(TT_name_re)
    # phy_TT_re <- Results_TT_re$phylo
    # mf_TT_re <- Results_TT_re$meanFitness[1:length(phy_TT_re$tip.label)]
    # Fitness_true_re <- as.numeric(phy_TT_re$fitness$V1[as.numeric(phy_TT_re$tip.label)])
    # corr_true_re[jj,ii-10] <- cor(mf_TT_re,Fitness_true_re,method='spearman')
    
    
    TT_name_ori <- paste0("Results/non-spatial simulation results/SIM_", batch_num,"/P_fitness.out.rds")
    Results_TT <- readRDS(TT_name_ori)
    phy_TT <- Results_TT$phylo
    mf_TT <- Results_TT$meanFitness[1:length(phy_TT$tip.label)]
    Fitness_true <- as.numeric(phy_TT$fitness$V1[as.numeric(phy_TT$tip.label)])
    corr_true_ori[ii] <- cor(mf_TT, Fitness_true, method = "spearman")
    
    dir_name_sub <- paste0("Results/non-spatial simulation results/SIM_", batch_num,"/Subsampling/")
  TT_name_sub <- paste0(dir_name_sub,"P_fitness-",case_num,"-",batch_num,"_",sample_num,".out.rds")
    Results_TT_sub <- readRDS(TT_name_sub)
    phy_TT_sub <- Results_TT_sub$phylo
    index_sub <- as.numeric(phy_TT_sub$tip.label)
    mf_TT_sub <- Results_TT_sub$meanFitness[1:length(phy_TT_sub$tip.label)]
    
    corr_ori_sub[jj,ii] <- cor(mf_TT_sub,mf_TT[index_sub],method='spearman')
    
    
    
    
  }
}

```


### Visualization (D plots)

```{r}

# re_df <- data.frame(
#   idx = 1:50,
#   Acc = apply(corr_true_re,2,mean),
#   re_var = apply(corr_true_re,2,var)
# )
# 
# cor_test <- cor.test(re_df$re_var, re_df$Acc)
# label_text <- paste0(
#   "Pearson = ", round(cor_test$estimate, 2),
#   "\nP = ", signif(cor_test$p.value, 2)
# )
# 
# p_re <- ggplot(re_df, aes(x = Acc, y = re_var)) +
#   geom_point(size = 3) +
#   geom_smooth(method = "lm", se = TRUE) +
#   annotate(
#     "text",
#     x = max(re_df$Acc) * 0.8,
#     y = max(re_df$re_var) * 0.9,
#     label = label_text,
#     size = 5
#   ) +
#   labs(
#     y = "Resampling Variance",
#     x = "Prediction Accuracy (Spearman)"
#   ) +
#   theme_bw()+
#   theme(
#     axis.text = element_text(size = 14),
#     plot.title = element_text(size = 18, face = "bold")
#   )


sub_df <- data.frame(
  idx = 1:50,
  Acc = corr_true_ori,
  sub_var = apply(corr_ori_sub,2,var)
)

sub_df$color <- ifelse(plot_df$idx == 24, "#1F78B4",
                        ifelse(plot_df$idx == 35, "#A13F55","black"))

cor_test <- cor.test(sub_df$sub_var, sub_df$Acc)
label_text <- paste0(
  "Pearson = ", round(cor_test$estimate, 2),
  "\nP = ", signif(cor_test$p.value, 2)
)

p_sub <- ggplot(sub_df, aes(x = Acc, y = sub_var)) +
  geom_point(aes(color = color), size = 2.5)+
  scale_color_identity()+
  geom_smooth(method = "lm", se = TRUE) +
  annotate(
    "text",
    x = max(sub_df$Acc) * 0.8,
    y = max(sub_df$sub_var) * 0.9,
    label = label_text,
    size = 5
  ) +
  labs(
    y = "Subsampling Variance",
    x = "Prediction Accuracy (Spearman)"
  ) +
  theme_bw()+
  theme(
    axis.text = element_text(size = 14),
    plot.title = element_text(size = 18, face = "bold")
  )




```

#### Supplementary plots


We collected all the inference results (TT tree) from 50 simulated tumor and store their correlation with true fitness as comparative plots in "MATLAB/Simulation/Results/Figures/SIM_plots".

```{r}
#i <- 35
for (i in 1:50){
  case_num <- "8"
  batch_num <- as.character(i)
  
  dir_name <- paste0("MATLAB/Simulation/Results/SIM_",case_num,"_",batch_num)
  TT_name <- paste0(dir_name,"/P_fitness-",case_num,"-",batch_num,"_TS1.out.rds")
  Results_TT <- readRDS(TT_name)
  
  phy_TT <- readRDS(paste0(dir_name,"/Tree-",case_num,"-",batch_num,".rds"))
  Fitness_true    <- as.numeric(phy_TT$fitness$V1)
  mf_TT <- Results_TT$meanFitness[1:length(phy_TT$tip.label)]
  
  
  Tf_ring <- tibble(
    label = phy_TT$tip.label,
    relative_fitness = Fitness_true
  )
  
  
  TT_ring <- tibble(
    label = phy_TT$tip.label,
    relative_fitness = mf_TT
  )
  
  
  
  
  p_Tf <- ggtree(phy_TT, layout = 'circular')+ labs(caption = "True relative fitness") +
    theme(legend.position = "right", plot.caption = element_text(size = 12, hjust = 0.5), legend.key.size = unit(0.2, "cm")) + 
    geom_fruit(
        data = Tf_ring,
        geom = geom_bar,
        mapping = aes(y = label, x = relative_fitness, fill = relative_fitness),
        orientation = "y",
        stat = "identity",
        width = 0.3
      ) +
    scale_fill_gradient(low = "#1F78B4", high = "darkblue") + guides(fill = guide_legend(order = 1), color = guide_legend(order = 2))
  
  

  
  
  p_TT <- ggtree(phy_TT, layout = 'circular')+labs(caption = "TT Inferred relative fitness") +
    theme(legend.position = "right", plot.caption = element_text(size = 12, hjust = 0.5), legend.key.size = unit(0.2, "cm")) + 
    geom_fruit(
        data = TT_ring,
        geom = geom_bar,
        mapping = aes(y = label, x = relative_fitness, fill = relative_fitness),
        orientation = "y",
        stat = "identity",
        width = 0.3
      ) +
    scale_fill_gradient(low = "#1F78B4", high = "darkblue")+ guides(fill = guide_legend(order = 1), color = guide_legend(order = 2))
  
  
  p <- (p_Tf|p_TT) + plot_annotation(title = paste0("Spearman Correlation: ", format(round(corr_true_ori[i], 4), nsmall = 4)))&
  theme(plot.title = element_text(size = 18, face = "bold", hjust = 0.5))


  ggsave(paste0("MATLAB/Simulation/Figures/SIM_plots/SIM-8-",batch_num,"-fitness.pdf"), plot = p, width = 7, height = 5)
}

```

We examine the accuracy of fitness prediction measured by mean relative error for each synthesis tumor.

```{r}
p1 <- ggplot(data.frame(x = acc), aes(x)) + geom_histogram(binwidth = 0.01, fill = "#1F78B4", color = "white") + scale_x_continuous(
    name = "Average Relative Absolute Error (%)",
    limits = c(0,0.1),
    breaks = c( 0, 0.02, 0.04, 0.06, 0.08, 0.1),
labels = c(
  "0%", 
  "2%", 
  "4%", 
  "6%", 
  "8%", 
  "10%"
)
  ) + labs(title = "Accuracy across 50 tumors", y = "Frequency")

#resample_corr <- corr_true_ori[20]

#p2 <- ggplot(data.frame(x = corr_true_sub_ori), aes(x)) + geom_histogram(binwidth = 0.025, fill = "#A13F55", color = "white") + scale_x_continuous(
#    name = "Spearman correlation significancy",
#    limits = c(-0.05,1),
#    breaks = c( 0, 0.2, 0.4, 0.6, 0.8, 1),
#labels = c(
#  "0\n(None)", 
#  "0.2\n(weak+)", 
#  "0.4\n(moderate+)", 
#  "0.6\n(strong+)", 
#  "0.8\n(very strong+)", 
#  "1\n(perfect+)"
#)) + labs(title = "Robustness of ranking under resampling", y = "Frequency")  + theme_minimal()

#p <- p1/p2

#ggsave("/projects/standard/ruping/shared/wu000766/Fitness_ranking/MATLAB/Simulation/Figures/SIM_robustness.svg", plot = p, width = 6, height = 8)

```


We collected the inference accuracy from all 100 re-sampled TT trees for tumor #20.

```{r}
corr_true_sub_ori <- rep(0,100)

for (ii in 1:100){
  case_num <- "8"
  batch_num <- "20"
  sample_num <- as.character(ii)
  dir_name <- paste0("MATLAB/Simulation/Results/SIM_",case_num,"_",batch_num,"/Resampling/")
  TT_name <- paste0(dir_name,"P_fitness-",case_num,"-",batch_num,"_",sample_num,".out.rds")
  Results_TT <- readRDS(TT_name)
  
  phy_ori <- readRDS(paste0(dir_name,"Tree-",case_num,"-",batch_num,"_",sample_num,".rds"))
  Fitness_true    <- as.numeric(phy_ori$fitness$V1)
  mf_TT <- Results_TT$meanFitness[1:length(phy_ori$tip.label)]
  corr_true_sub_ori[ii] <- cor(mf_TT,Fitness_true,method='spearman')
}


p_result <- ggplot(data.frame(x = corr_true_sub_ori), aes(x)) + geom_histogram(binwidth = 0.025, fill = "#A13F55", color = "white") + scale_x_continuous(
    name = "Spearman correlation significancy",
    limits = c(-0.05,1),
    breaks = c( 0, 0.2, 0.4, 0.6, 0.8, 1),
labels = c(
  "0\n(None)", 
  "0.2\n(weak+)", 
  "0.4\n(moderate+)", 
  "0.6\n(strong+)", 
  "0.8\n(very strong+)", 
  "1\n(perfect+)"
)) + labs(title = "Robustness of ranking under resampling", y = "Frequency")  + theme_minimal()

```


```{r}


p1 <- ggplot(data.frame(x = corr_true_ori), aes(x)) + geom_histogram(binwidth = 0.05, fill = "#1F78B4", color = "white") + scale_x_continuous(
    name = "Spearman correlation significancy",
    limits = c(-0.05,1),
    breaks = c( 0, 0.2, 0.4, 0.6, 0.8, 1),
labels = c(
  "0\n(None)", 
  "0.2\n(weak+)", 
  "0.4\n(moderate+)", 
  "0.6\n(strong+)", 
  "0.8\n(very strong+)", 
  "1\n(perfect+)"
)
  ) + labs(title = "Robustness of ranking", y = "Frequency")

resample_corr <- corr_true_ori[20]

p2 <- ggplot(data.frame(x = corr_true_sub_ori), aes(x)) + geom_histogram(binwidth = 0.025, fill = "#A13F55", color = "white") + scale_x_continuous(
    name = "Spearman correlation significancy",
    limits = c(-0.05,1),
    breaks = c( 0, 0.2, 0.4, 0.6, 0.8, 1),
labels = c(
  "0\n(None)", 
  "0.2\n(weak+)", 
  "0.4\n(moderate+)", 
  "0.6\n(strong+)", 
  "0.8\n(very strong+)", 
  "1\n(perfect+)"
)) + labs(title = "Robustness of ranking under resampling", y = "Frequency")  + theme_minimal()

p <- p1/p2

#ggsave("/projects/standard/ruping/shared/wu000766/Fitness_ranking/MATLAB/Simulation/Figures/SIM_robustness.svg", plot = p, width = 6, height = 8)


```

## Figure 3(In silico parameter robustness)

### Collect Panel A data

```{r}
param_cor_df <- data.frame(
    Ori = rep(0,50),
    rho001 = rep(0,50),
    rho00005= rep(0,50),
    TS6    = rep(0,50),
    d15    = rep(0,50),
    Type4li = rep(0,50),
    Type4log = rep(0,50),
    mu001  = rep(0,50),
    delta1 = rep(0,50)
)


type_num_hist <- rep(0,50)

for (case_i in 1:length(colnames(param_cor_df))){
  case_name <- colnames(param_cor_df)[case_i]
  case_num <- "8"
  for (ii in 1:50){
    
    batch_num <- as.character(ii)
    dir_name <- paste0("Results/non-spatial simulation results/SIM_",batch_num)
    if (case_i == 1){
      TT_name <- paste0(dir_name,"/P_fitness.out.rds")
    }else{
      TT_name <- paste0(dir_name,"/P_fitness_", case_name,".out.rds")
    }
    
    Results_TT <- readRDS(TT_name)
    
    phy_ori <- readRDS(paste0(dir_name,"/Tree.rds"))
    Fitness_true    <- as.numeric(phy_ori$fitness$V1)
    type_num_hist[ii] <- dim(table(Fitness_true))
    mf_TT <- Results_TT$meanFitness[1:length(phy_ori$tip.label)]
    param_cor_df[ii,case_i]<- cor(mf_TT,Fitness_true,method='spearman')
  }
}

```

### Panel A

```{r}


#col_names  <- c("Ori","2 Types", "4 Types","6 Types", "7 Types", expression(paste(lambda,"+")), expression(paste(mu, "+")), expression(paste(rho, "-")),expression(paste(rho,"+")), "Range -", "Range +","Time scale -", "Time scale +")
#svg("MATLAB/Simulation/Figures/Parameter's effect.svg", width = 15, height = 5)
#png("MATLAB/Simulation/Figures/Parameter's effect.png", width = 1000, height = 500)
boxplot(param_cor_df, ylab = "Spearman Correlation")
#dev.off()

```



### Panel B and C

In panel B and C, we compare the fitness density under all different scenario ("p_u","mp_ht_u") with the original inference ("p","mp_ht") and True Fitness ("p_t","mp_FT_ht").
The fitness plots are start with "p", and the misspecified fitness in Panel B is "p_u".
The variable "U_case" controls the scenario, including
1. TS2
2. TS4
3. TS6
4. TS8
5. d15
6. delta1
7. mu001
8. rho00005
9. rho001
10. Type4log    
11. Type4li
When plotting the mp_ht_u, it is necessary to adjust the column names according to the experimental design described in the manuscript as the number of fitness types is reduced to 4.


```{r}
# case_num <- "8li"
batch_num <- "38"
U_case <- "mu001"
dir_name <- paste0("Results/non-spatial simulation results/SIM_",batch_num,"/")
results_file <- paste0(dir_name,"P_fitness.out.rds")
Results <- readRDS(results_file)
tree <- Results$phylo
meanFitness <- Results$meanFitness[1:length(tree$tip.label)]
mp <- Results$marginalProb[1:length(tree$tip.label),]
colnames(mp) <- as.character(seq(0.2,0.8, length.out = 8))
hc_tree <- as.hclust(tree)
mp_ht <- Heatmap(
  mp,
  cluster_rows = hc_tree,
  cluster_columns = FALSE,
  col = c("#deebf7","#3182bd"), # This color can be used if necessary
  column_names_rot = 90,
  border = TRUE
)



Fitness_true    <- as.numeric(tree$fitness$V1)
mp_FT <- matrix(0, nrow = length(Fitness_true), ncol = 8)
mp_FT[cbind(seq_len(length(Fitness_true)), Fitness_true)] <- 1
colnames(mp_FT) <- as.character(seq(0.2,0.8, length.out = 8))
mp_FT_ht <- Heatmap(
  mp_FT,
  cluster_rows = hc_tree,
  cluster_columns = FALSE,
  col = c("#deebf7","#3182bd"),
  column_gap = unit(2,"mm"),
  column_names_rot = 90,
  border = TRUE
)

#svg("MATLAB/Simulation/Figures/(Misspecified Inference) True Fitness.svg", width = 4, height = 6)
#mp_FT_ht
#dev.off()



ring_t <- tibble(
  label = tree$tip.label,
  fitness = Fitness_true
)
  


  
  p_t <- ggtree(tree, layout = 'circular')+ ggtitle(paste0("True Fitness")) +
    labs(fill = "Fitness") +
    theme(legend.position = "bottom", 
          plot.title = element_text(size = 12, hjust = 0.5, vjust = -1, face = "bold"), 
          legend.key.size = unit(0.2, "cm"),
          legend.title.position = 'top') + 
    geom_fruit(
        data = ring_t,
        geom = geom_bar,
        mapping = aes(y = label, x = fitness, fill = fitness),
        orientation = "y",
        stat = "identity",
        width = 0.3
      ) +
    scale_fill_gradient(low = "#93C477", high = "#1F78B4") + 
    guides(fill = guide_colorbar(order = 1,
                                 barwidth = unit(6, "cm")))




  
ring <- tibble(
  label = tree$tip.label,
  fitness = meanFitness
)
  


  
  p <- ggtree(tree, layout = 'circular')+ ggtitle(paste0("Valid Inference: (\u03C1: ", format(round(cor(Fitness_true,meanFitness,method = 'spearman'), 4), nsmall = 4), ")")) +
    labs(fill = "Fitness") +
    theme(legend.position = "bottom", 
          plot.title = element_text(size = 12, hjust = 0.5, vjust = -1, face = "bold"), 
          legend.key.size = unit(0.2, "cm"),
          legend.title.position = 'top') + 
    geom_fruit(
        data = ring,
        geom = geom_bar,
        mapping = aes(y = label, x = fitness, fill = fitness),
        orientation = "y",
        stat = "identity",
        width = 0.3
      ) +
    scale_fill_gradient(low = "#93C477", high = "#1F78B4") + 
    guides(fill = guide_colorbar(order = 1,
                                 barwidth = unit(6, "cm")))
  
  #ggsave(paste0("Andrew2025/Results/Figures/OV-",OV_num,"_Fitness_c.svg"),plot = p, width = 6, height = 8)
  
  
  

results_file_u <- paste0(dir_name,"P_fitness_", U_case,".out.rds")
Results_u <- readRDS(results_file_u)
tree_u <- Results_u$phylo
meanFitness_u <- Results_u$meanFitness[1:length(tree$tip.label)]
mp_u <- Results_u$marginalProb[1:length(tree$tip.label),]
hc_tree_u <- as.hclust(tree_u)
b_rates <- 0.2*1.2^(0:7)
colnames(mp_u) <- as.character(0.2*1.2^(0:7))
mp_ht_u <- Heatmap(
  mp_u,
  cluster_rows = hc_tree,
  cluster_columns = FALSE,
  col = c("#deebf7","#3182bd"),
  column_names_rot = 90,
  border = TRUE
)

km_HL <- kmeans(meanFitness_u, centers = 3, nstart = 10)
  
ring_u <- tibble(
  label = tree_u$tip.label,
  fitness = meanFitness_u
)
  
  

  
  p_u <- ggtree(tree_u, layout = 'circular')+ ggtitle(paste0("Misspecified Inference: (\u03C1: ", format(round(cor(Fitness_true,meanFitness_u,method = 'spearman'), 4), nsmall = 4), ")")) +
    labs(fill = "Fitness") +
    theme(legend.position = "bottom", 
          plot.title = element_text(size = 12, hjust = 0.5, vjust = -1, face = "bold"), 
          legend.key.size = unit(0.2, "cm"),
          legend.title.position = 'top') + 
    geom_fruit(
        data = ring_u,
        geom = geom_bar,
        mapping = aes(y = label, x = fitness, fill = fitness),
        orientation = "y",
        stat = "identity",
        width = 0.3
      ) +
    scale_fill_gradient(low = "#93C477", high = "#1F78B4") + 
    guides(fill = guide_colorbar(order = 1,
                                 barwidth = unit(6, "cm")))

```






### Detailed study with Birth rates spacing

This section includes the analysis of reproducing the Figure S2.

```{r}
param_cor_df <- data.frame(
    Ori = rep(0,50),
    TS4Type4li = rep(0,50),
    TS4Type4log = rep(0,50)
)

True_var <- rep(0,50)

for (case_i in 1:length(colnames(param_cor_df))){
  case_name <- colnames(param_cor_df)[case_i]
  case_num <- "8"
  for (ii in 1:50){
    
    batch_num <- as.character(ii)
    dir_name <- paste0("MATLAB/Simulation/Results/SIM_",case_num,"_",batch_num)
    if (case_i == 1){
    TT_name <- paste0(dir_name,"/P_fitness-",case_num,"-",batch_num,"_TS4.out.rds")     }
    else{
    TT_name <- paste0(dir_name,"/P_fitness-",case_num,"-",batch_num,"_",case_name,".out.rds")
    }
    Results_TT <- readRDS(TT_name)
    
    phy_ori <- readRDS(paste0(dir_name,"/Tree-",case_num,"-",batch_num,".rds"))
    Fitness_true    <- as.numeric(phy_ori$fitness$V1)
    True_var[ii] <- sd(Fitness_true)
    mf_TT <- Results_TT$meanFitness[1:length(phy_ori$tip.label)]
    param_cor_df[ii,case_i]<- cor(mf_TT,Fitness_true,method="spearman")
  }
}




```


```{r}
param_cor_li_df <- data.frame(
    Ori_li = rep(0,50),
    TS4Type4li = rep(0,50),
    TS4Type4log = rep(0,50)
)

True_var_li <- rep(0,50)

for (case_i in 1:length(colnames(param_cor_li_df))){
  case_name <- colnames(param_cor_li_df)[case_i]
  case_num <- "8li"
  for (ii in 1:50){
    
    batch_num <- as.character(ii)
    dir_name <- paste0("MATLAB/Simulation/Results/SIM_",case_num,"_",batch_num)
    if (case_i == 1){
      TT_name <- paste0(dir_name,"/P_fitness-",case_num,"-",batch_num,"_TS4li.out.rds")
    }else{
      TT_name <- paste0(dir_name,"/P_fitness-",case_num,"-",batch_num,"_", case_name,".out.rds")
    }
    Results_TT <- readRDS(TT_name)
    
    phy_ori <- readRDS(paste0(dir_name,"/Tree-",case_num,"-",batch_num,".rds"))
    Fitness_true    <- as.numeric(phy_ori$fitness$V1)
    True_var_li[ii] <- sd(Fitness_true)
    mf_TT <- Results_TT$meanFitness[1:length(phy_ori$tip.label)]
    param_cor_li_df[ii,case_i]<- cor(mf_TT,Fitness_true,method='spearman')
  }
}

```

## Figure 4 (In silico WGD robutness) Panel B, Panel D, and Panel F

### Collecting correlation  (non spatial) Data for panel F

```{r}
num_sim <- 50

corr_true_ori <- rep(0,num_sim)
corr_true_wgd <- rep(0,num_sim)
corr_true_dm  <- rep(0,num_sim)
corr_ori_wgd  <- rep(0,num_sim)


for (i in 1:num_sim){
  case_num <- "8"
  batch_num <- as.character(i)
  dir_name <- paste0("Results/non-spatial simulation results/SIM_WGD_",batch_num)
  TT_name <- paste0(dir_name,"/P_fitness_TT.out.rds")
  Results_TT <- readRDS(TT_name)
  DM_name <- paste0(dir_name,"/P_fitness_DM.out.rds")
  Results_DM <- readRDS(DM_name)
  WGD_name <- paste0(dir_name,"/P_fitness_WGD.out.rds")
  Results_WGD <- readRDS(WGD_name)
  
  phy_ori <- readRDS(paste0(dir_name,"/Tree.rds"))
  Fitness_true    <- as.numeric(phy_ori$fitness$V1)
  
  mf_TT <- Results_TT$meanFitness[1:length(phy_ori$tip.label)]
  mf_DM <- Results_DM$meanFitness[1:length(phy_ori$tip.label)]
  mf_WGD <- Results_WGD$meanFitness[1:length(phy_ori$tip.label)]
  
  corr_true_ori[i]<- cor(mf_TT,Fitness_true,method='spearman')
  corr_true_dm[i]<- cor(mf_DM,Fitness_true,method='spearman')
  corr_true_wgd[i]<- cor(mf_WGD,Fitness_true,method='spearman')
  corr_ori_wgd[i] <- cor(mf_TT,mf_WGD,method = 'spearman')
}

#writeMat("Corr_results.mat",TT = corr_true_ori, DM = corr_true_dm, WGD = corr_true_wgd)
```


### Panel B

```{r}
i <- 20
#for (i in 1:50){
  case_num <- "8"
  batch_num <- as.character(i)
  
  dir_name <- paste0("Results/non-spatial simulation results/SIM_WGD_",batch_num)
  TT_name <- paste0(dir_name,"/P_fitness_TT.out.rds")
  Results_TT <- readRDS(TT_name)
  DM_name <- paste0(dir_name,"/P_fitness_DM.out.rds")
  Results_DM <- readRDS(DM_name)
  WGD_name <- paste0(dir_name,"/P_fitness_WGD.out.rds")
  Results_WGD <- readRDS(WGD_name)
  tip_wgd <- unlist(read.csv(paste0(dir_name,'/WGD.txt'),header = FALSE),use.names = FALSE)-1
  
  phy_ori <- readRDS(paste0(dir_name,"/Tree.rds"))
  Fitness_true    <- as.numeric(phy_ori$fitness$V1)
  mf_TT <- Results_TT$meanFitness[1:length(phy_ori$tip.label)]
  mf_DM <- Results_DM$meanFitness[1:length(phy_ori$tip.label)]
  mf_WGD <- Results_WGD$meanFitness[1:length(phy_ori$tip.label)]
  phy_DM <- Results_DM$phylo
  phy_WGD <- Results_WGD$phylo
  
  wgd_groups <- ifelse(tip_wgd == 0, "WGD -","WGD +")
  
  groups <- data.frame(
    label = phy_ori$tip.label,
    group = wgd_groups
  )
  otu_groups <- split(groups$label, groups$group)
  tree_grouped <- groupOTU(phy_ori, otu_groups)
  tree_DM_grouped <- groupOTU(phy_DM, otu_groups)
  tree_WGD_grouped <- groupOTU(phy_WGD, otu_groups)
  
  
  Tf_ring <- tibble(
    label = phy_ori$tip.label,
    fitness = Fitness_true
  )
  
  
  TT_ring <- tibble(
    label = phy_ori$tip.label,
    fitness = mf_TT
  )
  
  DM_ring <- tibble(
    label = phy_ori$tip.label,
    fitness = mf_DM
  )
  
  WGD_ring <- tibble(
    label = phy_ori$tip.label,
    fitness = mf_WGD
  )
  
  
  p_Tf <- ggtree(tree_grouped, aes(color=group), layout = 'circular', size = 0.3)+ 
    ggtitle("True ranking") +
    theme(plot.title = element_text(size = 14, vjust = -2.5,hjust = 0.5, face = "bold"), 
          legend.position = "bottom",
          legend.key.size = unit(0.3, "cm"),
          legend.title = element_text(size = 14, face = "bold"),
          legend.text = element_text(size = 12, face = "bold")
          ) + 
    geom_fruit(
        data = Tf_ring,
        geom = geom_bar,
        mapping = aes(y = label, x = fitness, fill = fitness),
        orientation = "y",
        stat = "identity",
        width = 0.3
      ) +
    scale_fill_gradient(low = "#93C477", high = "#1F78B4") + 
    guides(fill = guide_colorbar(order = 1,
                                 barwidth = unit(6, "cm"),
                                 barheight = unit(0.4,"cm"),
                                 title.position = "top"), 
           color = guide_legend(order = 2,
                                title.position = "top"))
  p_Tf <- p_Tf + scale_color_manual(values = c("WGD -" = "#435274", "WGD +" = "#A13F55"))
  
  

  
  
  p_TT <- ggtree(tree_grouped, aes(color=group), layout = 'circular', size = 0.3)+
    ggtitle(paste0("TT ranking (\u03C1: ",format(round(corr_true_ori[i], 4), nsmall = 4),")")) +
    theme(plot.title = element_text(size = 14, vjust = -2.5,hjust = 0.5, face = "bold"), 
          legend.position = "bottom",
          legend.key.size = unit(0.3, "cm"),
          legend.title = element_text(size = 14, face = "bold"),
          legend.text = element_text(size = 12, face = "bold")
          ) + 
    geom_fruit(
        data = TT_ring,
        geom = geom_bar,
        mapping = aes(y = label, x = fitness, fill = fitness),
        orientation = "y",
        stat = "identity",
        width = 0.3
      ) +
    scale_fill_gradient(low = "#93C477", high = "#1F78B4") + 
    guides(fill = guide_colorbar(order = 1,
                                 barwidth = unit(6, "cm"),
                                 barheight = unit(0.4,"cm"),
                                 title.position = "top"), 
           color = guide_legend(order = 2, 
                                title.position = "top"))
  p_TT <- p_TT + scale_color_manual(values = c("WGD -" = "#435274", "WGD +" = "#A13F55"))
  
  rng <- range(DM_ring$fitness, na.rm = TRUE)
  b_rng <- c(rng[1] + 0.05*diff(rng), rng[2] - 0.05*diff(rng))
  
  p_DM <- ggtree(tree_DM_grouped, aes(color=group), layout = 'circular', size = 0.3)+
    ggtitle(paste0("DM ranking (\u03C1: ",format(round(corr_true_dm[i], 4), nsmall = 4),")")) +
    theme(plot.title = element_text(size = 14, vjust = -2.5,hjust = 0.5, face = "bold"), 
          legend.position = "bottom",
          legend.key.size = unit(0.3, "cm"),
          legend.title = element_text(size = 14, face = "bold"),
          legend.text = element_text(size = 12, face = "bold")
          ) + 
    geom_fruit(
        data = DM_ring,
        geom = geom_bar,
        mapping = aes(y = label, x = fitness, fill = fitness),
        orientation = "y",
        stat = "identity",
        width = 0.3
      ) +
    scale_fill_gradient(low = "#93C477", high = "#1F78B4",
                        limits = rng,
                        breaks = b_rng,
                        labels = c("Low","High"),
                        name = "Fitness") + 
    guides(fill = guide_colorbar(order = 1,
                                 barwidth = unit(6, "cm"),
                                 barheight = unit(0.4,"cm"),
                                 title.position = "top"), 
           color = guide_legend(order = 2,
                                title.position = "top"))
  p_DM <- p_DM + scale_color_manual(values = c("WGD -" = "#435274", "WGD +" = "#A13F55"))
  
  p_WGD <- ggtree(tree_WGD_grouped, aes(color=group), layout = 'circular', size = 0.3)+
    ggtitle(paste0("WGD ranking (\u03C1: ",format(round(corr_true_wgd[i], 4), nsmall = 4),")")) +
    theme(plot.title = element_text(size = 14, vjust = -2.5,hjust = 0.5, face = "bold"), 
          legend.position = "bottom",
          legend.key.size = unit(0.3, "cm"),
          legend.title = element_text(size = 14, face = "bold"),
          legend.text = element_text(size = 12, face = "bold")
          ) + 
    geom_fruit(
        data = WGD_ring,
        geom = geom_bar,
        mapping = aes(y = label, x = fitness, fill = fitness),
        orientation = "y",
        stat = "identity",
        width = 0.3
      ) +
    scale_fill_gradient(low = "#93C477", high = "#1F78B4") + 
    guides(fill = guide_colorbar(order = 1,
                                 barwidth = unit(6, "cm"),
                                 barheight = unit(0.4,"cm"),
                                 title.position = "top"), 
           color = guide_legend(order = 2,
                                title.position = "top"))
  p_WGD <- p_WGD + scale_color_manual(values = c("WGD -" = "#435274", "WGD +" = "#A13F55"))
  
  
  
  
  shard_legend <- cowplot::get_legend(p_DM)
  
  p_Tf_n <- p_Tf + theme(legend.position = "none")
  p_TT_n <- p_TT + theme(legend.position = "none")
  p_DM_n <- p_DM + theme(legend.position = "none")
  p_WGD_n <- p_WGD + theme(legend.position = "none")
  
  

  title_grob <- textGrob(
    "Non-Spatial Simulation Analysis",
    gp = gpar(fontface = "bold", fontsize = 16),
    just = "center")
  
  p <- grid.arrange(
    arrangeGrob(p_Tf_n,p_TT_n,p_DM_n,p_WGD_n, nrow = 1), 
    shard_legend,
    title_grob,
    heights = c(0.5,0.1,0.1),
    ncol = 1
  )

  
```


### Panel D

```{r}
case_num <- "8"
batch_num <- as.character(20)

  

dir_name     <- paste0("Results/non-spatial simulation results/SIM_WGD_", batch_num,"/")
    DM_name <- paste0(dir_name,"Tree-DM.rds")
    TT_name <- paste0(dir_name,"Tree.rds") # Here we use true total time tree to determine the true overall time horizon
    
    
    
    
    #Ori_tree <- read.tree(name)
    #Ori_tree <- readRDS(name)
    DM_tree  <- readRDS(DM_name)
    TT_tree  <- readRDS(TT_name)
    WGD_file <- paste0(dir_name,"ultra-tree.csv")
    
    #discrete_tree <- readRDS(discrete_file)
    #WGD_R_tree <- readRDS(WGD_R_file)
    WGD_Results <- read.csv(WGD_file, header = FALSE)
    WGD_Results <- WGD_Results[,1]
    WGD_tree <- DM_tree
    e1 <- WGD_tree$edge[,1]
    e2 <- WGD_tree$edge[,2]
    li <- length(WGD_Results)
    
    WGD_ages <- c(rep(0,li),1,WGD_Results[3:length(WGD_Results)])
    WGD_tree$edge.length <- WGD_ages[e1] - WGD_ages[e2]
    DM_tree$node_wgd <- unlist(read.csv(paste0(dir_name,'WGD.txt'),header = FALSE),use.names = FALSE)
    DM_tree$node_wgd <- c(DM_tree$node_wgd-1,rep(-1,DM_tree$Nnode))
    DM_tree <- label_internal_nodes(DM_tree)

y <- DM_tree$edge.length[DM_tree$edge_wgd == 1]
t <- WGD_tree$edge.length[DM_tree$edge_wgd == 1]


WGD_rates <- WGD_Results[2]

tau <- WGD_rates*t
n   <- length(y)

k_max <- max(y)
k <- 0:k_max
O <- tabulate(y+1, nbins = k_max+1)
E <- sapply(k, function(kk){
  sum(dpois(kk,lambda = tau))
})


obs_stat <- 2 * sum(
  ifelse(y == 0,
         0,
         y * log(y / tau)) -
    (y - tau)
)

B <- 1000
sim_stat <- numeric(B)
for (b in 1:B) {
  sim_branch <- rpois(length(tau), tau)
  
  sim_stat[b] <- 2 * sum(
    ifelse(sim_branch == 0,
           0,
           sim_branch * log(sim_branch / tau)) -
      (sim_branch -  tau)
  )
}

p_sim_nWGD <- mean(sim_stat >= obs_stat)
p_sim_nWGD



#svg("MATLAB/Simulation/Figures/SIM_20_rooto.svg", width = 9, height = 4)

par(mfrow = c(1,2), oma = c(0,0,3,0))

y <- DM_tree$edge.length[DM_tree$edge_wgd == 0]
t <- WGD_tree$edge.length[DM_tree$edge_wgd == 0]


WGD_rates <- WGD_Results[1]

tau <- WGD_rates*t
n   <- length(y)

k_max <- max(y)
k <- 0:k_max
O <- tabulate(y+1, nbins = k_max+1)
E <- sapply(k, function(kk){
  sum(dpois(kk,lambda = tau))
})

sqrt_exp <- sqrt(E)
sqrt_obs <- sqrt(O)

signs <- sign(sqrt_exp - sqrt_obs)
sign_nWGD <- runs.test(as.factor(signs))

plot(k, sqrt_exp,
     type = "n",
     ylim = range(c(sqrt_exp, sqrt_exp - sqrt_obs)),
     xlab = "WGD- Branch Length",
     ylab = "\u221AFrequency",
     font.lab = 2)

# expected curve
lines(k, sqrt_exp, lwd = 2)

# hanging bars
segments(x0 = k,
         y0 = sqrt_exp,
         x1 = k,
         y1 = sqrt_exp - sqrt_obs,
         lwd = 3,
         col = "#435274") # WGD: #F8766D, non-WGD: #00BFC4

abline(h = 0, lty = 2)

legend("topright",
       legend = c("Expected","Observed"),
       col = c("black","#435274"),
       lty = 1,
       lwd = 3,
       bty = "n")


y <- DM_tree$edge.length[DM_tree$edge_wgd == 1]
t <- WGD_tree$edge.length[DM_tree$edge_wgd == 1]


WGD_rates <- WGD_Results[2]

tau <- WGD_rates*t
n   <- length(y)

k_max <- max(y)
k <- 0:k_max
O <- tabulate(y+1, nbins = k_max+1)
E <- sapply(k, function(kk){
  sum(dpois(kk,lambda = tau))
})

sqrt_exp <- sqrt(E)
sqrt_obs <- sqrt(O)

obs_stat <- 2 * sum(
  ifelse(y == 0,
         0,
         y * log(y / tau)) -
    (y - tau)
)

B <- 1000
sim_stat <- numeric(B)
for (b in 1:B) {
  sim_branch <- rpois(length(tau), tau)
  
  sim_stat[b] <- 2 * sum(
    ifelse(sim_branch == 0,
           0,
           sim_branch * log(sim_branch / tau)) -
      (sim_branch -  tau)
  )
}

p_sim_WGD <- mean(sim_stat >= obs_stat)
p_sim_WGD


plot(k, sqrt_exp,
     type = "n",
     ylim = range(c(sqrt_exp, sqrt_exp - sqrt_obs)),
     xlab = "WGD+ Branch length",
     ylab = "",
     font.lab = 2)

# expected curve
lines(k, sqrt_exp, lwd = 2)

# hanging bars
segments(x0 = k,
         y0 = sqrt_exp,
         x1 = k,
         y1 = sqrt_exp - sqrt_obs,
         lwd = 3,
         col = "#A13F55") # WGD: #F8766D, non-WGD: #00BFC4

abline(h = 0, lty = 2)


mtext("Hanging Rootogram for Non-Spatial Simulation",
      side = 3, outer = TRUE, line = 1, cex = 1.3, font = 2)

legend("topright",
       legend = c("Expected","Observed"),
       col = c("black","#A13F55"),
       lty = 1,
       lwd = 3,
       bty = "n")




#dev.off()




signs <- sign(sqrt_exp - sqrt_obs)
sign_WGD <- runs.test(as.factor(signs))
```







